                                                         5
                                          Chapter                 




          Modelling of concrete at high tempera-
          ture: thermodynamics, heat and mass 
          transfer 




 

5.1 Introduction 
In this chapter different aspects concerning the behaviour of concrete 
subjected to high temperatures and pressures will be analysed, first from 
thermodynamic physical point of view, then studying the inner 
microstructure of the material and its changes in such extreme conditions. 
The experimental and qualitative results of european research project 
BRITE "Hiteco" and their employment, in order to adapt the presented 
model for the study of innovative material such as high performance and 
ultra high performance concrete, will be shown. 
The developments of high performance (HPC) and ultra-high performance 
(UHPC) concrete open new possibilities for the design of concrete structures, 
both for high rise buildings and more slender sections in concrete structures 
such as bridges. 
Knowledge of the behaviour of these concretes at high temperatures is, 
however, limited. A major factor, particularly with dense high performance 
concrete, is the increased tendency to experience explosive spalling. 
One of the aims of this work is the use of the proposed model for analysing 
this complex material for a better understanding of the behaviour of 
concrete structures in situations involving high temperatures such as fire 
and nuclear rectors. 
In the framework of above mentioned "Hiteco" programme, a software 
implementing the model was developed. This fully coupled code is designed 


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




to predict the hygral-thermal-structural state, and spalling, of heated 
concrete structures. It considers concrete as multiphase material consisting 
of a solid phase, two gas phases and three water phases. 
Refinements, such as the effect of damage on permeability and non-
linearities due to temperature, are included and phase changes such as 
hydration-dehydration, evaporation-condensation adsorption-desorption are 
considered. 
Hence, concrete is treated as multiphase system [1]-[11] where the voids of 
the skeleton are filled partly with liquid- and partly with a gas-phase. The 
liquid phase consists of bound water (or adsorbed water), which is present in 
the whole range of water contents of the medium, and capillary water (or 
free water), which appears when water content exceeds so called solid 
saturation point Sssp [12], i.e. the upper limit of the hygroscopic region of 
moisture content, Figure 5.1. 
In fact, as it is clearly visible in Figure 5.1, there are two main regions 
depending on the state (free or bound) of the water present in the REV. 
Region I: in this region, called the nonhygroscopic region, free water exists 
within the REV (0<S<1) and the solid phase is fully saturated with bound 
water. The transition point between the region I and the region II is reached 
when all free water has been removed from the pores and the solid phase 
remains saturated with bound water. This point is known as the solid 
saturation point (S=0, or W=Wssp). 
 
                   Non-Hygroscopic region                                 Hygroscopic region 
                                 I                                                 II 
                                                                     





                                                g              s

                                                         w
                                                                         W=Wssp
                                                                         S=0


 
      Funicolar region Ia               Pendular region Ib 


           Figure 5.1 hygroscopic region, non-hygroscopic region and solid 
           saturation point in a porous media. 



Some authors divide the "nonhygroscopic region" into two parts by using the 
concept of the irreducible saturation point (S=Sirr, or W=Wirr) with the view 



 168


                                           Introduction 




that above that level the liquid phase is continuous (region Ia called 
funicolar region) and discontinuous below it (region Ib called pendular 
region) 





                                                                     s

                                 ws                             w
                              wgs





 

            Figure 5.2   liquid bridge between two particles in the pendular zone; 



Other authors prefer to assume that free water exists in only funicolar state 
above the solid saturation point, especially for hygroscopic media like 
concretes. 
Region II: in this region, called "hygroscopic region", only bound water 
remains in the REV. 
 Thus, in the governing equations for the hygroscopic region of moisture 
content, the liquid phase is the bound water: 

      for SSssp                        l=b,  h
                                                               =    h
                                                                                      
                                                   phase                  adsorp


while for the capillary region it is assumed that only capillary water can 
move, hence 

      for S>Sssp l=w, 
                                                  h
                                                               =         h
                                                                                 
                                                       phase              vap


The gas phase, i.e. moist air, is a mixture of dry air (non-condensable 
constituent) and water vapour (condensable gas), and is assumed to behave 
as an ideal gas. 
 




 169


      Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




Then it has been taken into account that hydration results in an increase of 
solid volume, thus also decrease of porosity according to  

          n                           m
                                                                
              dehydr                                 dehydr
                                                                
                               =                                 (5.1) 
                                                          s
               t                    t                      

It should be underlined, that in the mathematical model, equations (3.73) to 
(3.77), the terms containing the solid skeleton density are present. The 
latter must fulfil the solid mass conservation equation, 

            (1 - )                       s s              s
                                                                                        m
                          n         D                D n
                                                -               + (1 - n )       s       dehydr
                                                                              div v =                    (5.2) 
                     s                                                                             s
                                   Dt               Dt                                       

thus only two from the three material parameters describing the solid 
phase, i.e. n, s and m                                         are independent. 
                                                dehydr
  





 170


                 Physics and thermodynamics of fluids in concrete at high temperature 





5.2  Physics and thermodynamics of fluids in 
      concrete at high temperature 
Porous media are usually analysed from the macroscopic point of view 
accepting the continuum hypothesis of matter and then ignoring the 
discrete nature of the matter itself. This is the common approach used in 
thermodynamics. 
The underlying assumption is that at larger scales the macroscopic 
quantities can be identified as statistical averages of molecular quantities. 
When the equations governing the system are written in a local form, it is 
then implied that the elementary volume considered is large enough so that 
the thermodynamic properties (that are averaged molecular properties) are 
still meaningful. Moreover, to model a process in a continuum medium, 
where there are gradients of the thermodynamic coordinates (such as 
pressure or temperature), a local thermodynamic equilibrium hypothesis 
must be assumed (where local refers to the order of magnitude of the mean 
free path of the molecules). A rigorous assessment of such hypothesis 
requires to enter the realm of non-equilibrium thermodynamics and to 
evaluate the relaxation times and the molecular mean free path. In other 
words it is necessary to establish that the time scale of the variations of the 
macroscopic system are much longer than the time required to reach 
equilibrium locally and that in the length scale considered there is a number 
of molecules large enough so that the averaged molecular properties, i.e., 
thermodynamic coordinates, become meaningful. This is true when the time 
required for the redistribution of energy between the internal degrees of 
freedom of the molecules can be neglected (negligible bulk viscosity), such as 
an ideal gas, and when the medium is not too rarefied (i.e. Knudsen number 
>> 0.1). 
The porous medium poses some additional problems because it is a 
heterogeneous system. In fact, it can be considered as a multiphase system 
where the pores of a solid matrix are partly filled with liquid water and 
partly with a gaseous mixture of dry air and water vapour. To treat the 
porous medium as a system in which the properties are continuous 
functions of the space coordinates, some kind of larger scale volume 
averaging must be used (in addition to the statistical average at the 
molecular scale inside each phase, needed to define local thermodynamic 
properties). 
As shown in Chapter 2, this is can be done introducing the concept of 
representative elementary volume (REV). Obviously, this representative 
elementary volume has a length scale much grater than the molecular one 
previously mentioned and includes different components/phases that could 
have different mean values of the system coordinates (e.g., the mean 
temperature of the solid matrix could be different from the mean 
temperature of the fluid phase inside the REV, see Chapter 2). In principle, 
many averaging/homogenisation techniques take into account different 



 171


      Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




properties in different phases/components. But, usually, during the 
homogenisation process some terms are assumed to be negligible and some 
kind of integral is performed over the REV introducing a weighted average 
of the relevant variable. This leads to a redefinition of the related transport 
coefficients that incorporates, in more or less accurate fashion, the effects of 
heterogeneity. 
 As shown in Chapter 2, the general approach to the description of non-
equilibrium processes, including heat and mass transfer in porous medium, 
is to start from an appropriate set of conservation equations and to 
supplement them with thermodynamic equilibrium (or state) equations as 
well as constitutive equations. 
The equilibrium equations can be considered a generalisation of the 
equations of state and express the relationship existing between the 
thermodynamic coordinates and/or their variations at thermodynamic 
equilibrium. They then limit the number of independent coordinates needed 
to describe the state of the system (see Chapter 2). 

5.2.1  State equation of water 
The state equation of liquid water is rather complex [2],[13] and usually 
creates numerical problems. As a sufficient approximation of the 
dependence of the bulk density of liquid water upon its temperature and 
pressure, the following formula may be used [13],[14],  

                  w                   wo
                                          1  ( -  ) +  ( w                  wo
                                                                         -               )
                            =                     -
                                                      w    o     w p         p
                                                                                           (5.3) 

where wo= 999.84 kg/m3 is the water density at the reference temperature 
qo and pressure pwo,  
             1 
 =                                          is volume thermal expansion coefficient of the water, 
             v ( vw
 w                           )
             w                        p
                   1                  w
                             v 
      = -                                  
 w
                                                is the isothermal compression modulus of water 
                       w
                  v  p 
(w=4.310-9 Pa-1 [14]),  
and  w                      ( w
        v                    ) 1
                                       -
                  =                               is the water specific volume. The coefficient w changes in 
non-linear manner with temperature (e.g. w=0.6810-4K-1 at q=273.15 K and 
w=10.110-4K-1 at q=420 K [13]), thus the average value for the temperature 
range under interest should be used in calculations.  
The state equation (5.3), basically valid for the bulk (free) liquid water, is 
often used also for the description of the capillary and bound water, e.g. 
[15]-[18]. However, because of a very complex nature of interaction between 
the water and skeleton, its applicability is questionable, especially as far as 
pressure dependence is concerned (although capillary water is in traction, 
but its density is not expected to be lower than that for bulk water). 
Nevertheless, equation (5.3) may be used, taking into account that close to 



 172


                      Physics and thermodynamics of fluids in concrete at high temperature 




critical point of water a sharp increase of volume thermal expansion 
coefficient takes place and assuming water incompressibility, i.e. w=0. In 
Figure 5.3 a linear relationship with w=constant and w=4.310-9 Pa-1 is 
shown. This law is valid up to 200°C. 

                                                          1020

                      ]                                                                                                                                     100 kPa
                 3                                        1000                                                                                              200 kPa
                                                                                                                                                            500 kPa
                                                               980                                                                                          1000 kPa

                                                               960


                                                               940


                           Liquid water density  [kg/m         920


                                                               900
                                                                     273                                 373                     473                   573                 673

                                                                                                                         Temperature  [K]



          Figure 5.3  Water density: linear relationship eq. (5.3) with w=constant, 
          w=4.310-9 Pa-1 



Another relationship that can be used as state equation of water is a non-
linear expression of empirical nature due to Furbish [19]. Also in this case 
the incompressibility of water can be supposed. By assuming that  w
                                                                                                                                                                                    v  
remains constant, i.e. the liquid water compressibility tends to zero, 
     w
  0 : 

                                                          1               w
                                                                     
                                                                                                   1               w
                                                                                                               v
                                                                                                          
          w
           =                                                                       = -                                     (5.4) 
                                                          w               w                         w               w
                                                                    p
                                                                                                         
                                                                               H               v               p H

where  w
          v  is the molar volume of water and H is the main curvature of the 
liquid-vapour interface : 

          w
           = (                                                                           2                         3            4           5
                                                          b + b  + b  + b  + b  + b 
                                                          0               1          2                    3                 4           5         )
                                                                                                                                                                                 (5.5) 
               + (p - p                                                                   a + a  + a  + a  + a  + a 
                                                               w1              wrif )(                                           2           3              4         5
                                                                                               0               1            2           3              4         5         )

where: pw1=1e07 Pa and pwrif=2e07 Pa; the coefficients a and b in equation 
(5.5) are shown in table 5.1: 
 



 173


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




 
       a0                                           a1                 a2                  a3            a4              a5 
     4.89e07  -1.65e-09 1.86e-12  2.43e-13 -1.60e-15 3.37e-18 
       b0                                           b1                 b2                  b3            b4              b5 
     1.02e-03 -7-74e-01 8.77e-03 -9.21e-05 3.35e-07 -4.40e-10 


                                                         Table 5.1 coefficients of equation (5.4) 



The relationship of eq. (5.5) is reported in Figure 5.4 where is evident the 
dependence of water density on temperature and the thermal expansion of 
water close to critical point. 
 
                                                 1200

                   ]3 1000

                                                  800


                                                  600

                                                  400            Furbish formula
                                                                 Exper. values
                                                  200
                   Liquid water density [kg/m

                                                    0
                                                         0          100             200           300           400

                                                                              Temperature [°C]



              Figure 5.4 Water density: comparison between Furbish formula (5.5) 
              and experimental values 



5.2.2  Saturated vapour pressure 
The saturated water vapour pressure, i.e. the vapour pressure with flat 
surface (H=0), can be calculated through the well known Clasius-Clapeyron 
equation (see Chapter 2). Anyway, this relationship should be used carefully 
over extended temperature intervals, because enthalpy of vaporization is 
not constant but it is a function of temperature, i.e.  va
                                                                                                           h p =  v
                                                                                                                       h ap()  and 
specific volumes vw and vv, of liquid water and water vapour respectively, 
change with temperature, too (see Fig. 5.6 and 5.7). 
More accurate results can be obtained using empirical correlations such as 
Hyland-Wexler relationship (Figure 5.5): 




 174


                                      Physics and thermodynamics of fluids in concrete at high temperature 




                              gws
                         p           ( ) = exp[                                                   2              3
                                                           C /  + C + C  + C  + C  + C ln( )  (5.6) 
                                                            1           2         3           4              5              ]
                                                                                                                       6

where C1=5800.2206, C2=1.3914993, C3=-4.8640293.10-2, C4=4.1764768.10-5, 
C5=-1.4452093.10-8, C6=6.5459673. 
 


                         1,80E+06

                         1,60E+06

                         1,40E+06                                Hyland-Wexler

                                                                 Experimental values
                         1,20E+06

            a] 1,00E+06
              [P
     gws                 8,00E+05
                    P

                         6,00E+05

                         4,00E+05

                         2,00E+05

                         0,00E+00
                                       275 285 295 305 315 325 335 345 355 365 375 385 395 405 415 425 435 445 455 465

                                                                                       Temperature [K]



                          Figure 5.5 Saturation vapour pressure as function of temperature 
                          (Hyland-Wexler formula) in comparison with experimental values 



5.2.3  Thermodynamic equilibrium relationships 
It is well known that due to the curvature of the interface between the 
capillary- or bound-water and the gas phase inside the pores of the medium 
(considered as a capillary porous body), the equilibrium water vapour 
pressure pgw differs from the saturation pressure pgws and can be obtained 
from the Kelvin equation [20], 

                                                                                              c
                                                                                         p Mw 
                              gw            gws
                         p           = p           ()                gws
                                                           RH = p           () exp                   
                                                                                        -                (5.7) 
                                                                                              w
                                                                                          R 

where  RH is relative humidity, and the water vapour saturation pressure 
pgws, depending only upon temperature q, can be calculated from the 
Clausius-Clapeyron equation (see Chapter 2) or from empirical correlations, 
e.g. the formula proposed by Hyland and Wexler (5.6). 




 175


           Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




The radius of the curvature of the interface, Rc, being in the mechanical 
equilibrium state with water vapour, and corresponding to it pore radius, r, 
are given by the Young-Laplace equation [15],[20], 

                 c               wa
                                
               p =                        , (5.8) 
                                     c
                                R

where for spherical meniscus (capillary water) its curvature equals to 

                1               2  cos 
                           =                         (5.9) 
                      c
               R                 r - a
                                               t

while for cylindrical interface (bound water) it is  

                 1                   1
                           =                         (5.10) 
                R               r - t
                      c                   a


  is the surface tension of water-air interface, J is the wetting angle, 
     wa
usually assumed to be zero, hence cosJ=1, and ta is the thickness of the 
adsorbed water layer in the pore (see Fig. 5.20). 
As briefly mentioned in Chapter 2, it should be emphasised that equations 
(5.7) and (5.8) are valid both for the capillary and adsorbed water [20]. In 
the latter case the capillary pressure pc should be understood as the matric 
potential , being a thermodynamic potential referring to the bound water, 
which cannot be identified to the pressure [15] as it will be shown below. 
 
For a mixture of N components, the equilibrium between a liquid (l) and its 
vapour (v) can be expressed using the chemical potential µ as (supposing 
that chemical equilibrium exists): 

               µ = µ                                    i =                    N  (5.11) 
                 l                   v                         1,2,.......,
                                     i
                      i


or, since for a pure substance the chemical potential µ is equal to the molar 
Gibbs function g of each phase, the equilibrium can also be expressed: 

               g = g  (5.12) 
                 l              v

The differential of the chemical potential for a single phase of pure 
substance can be expressed as: 

               dg = 
                                - d + v dp + µdx - dL  (5.13) 

where  is the molar entropy, v  the molar volume, x the molar fraction and 
dL the amount of work made by the surface tension to extend the phase 
separation surface. The phase interface is assumed to have a constant area 




 176


                       Physics and thermodynamics of fluids in concrete at high temperature 




(i.e. dL=0) and for a pure substance the molar fraction x is always constant, 
then the variation of the molar Gibbs function is: 

       dg = 
                       - d + v dp  (5.14) 

It is important to highlight that the interfacial layer is not included in the 
analysis that means that the interfacial volume is negligible, i.e. V  0, 
                                                                                                     t
hence the Helmotz free energy of the interface tends to zero because there is 
only the work of surface formation whose variation is zero (as above 
mentioned, the interface is assumed to have a constant area), and the 
complete and ordinary chemical potential are equal (the derivatives of 
Helmotz free energy of the interfacial surface on numbers of moles are zero). 
Another hypothesis is that the meniscus mean radius of curvature is larger 
than the interfacial thickness, i.e. H-1> 100 otherwise the effect of 
curvature on surface tension is not more negligible (H is the curvature of the 
interface). 
Since  dµ = dµ  and thermodynamic equilibrium requires thermal 
                  l                 v
equilibrium, i.e., equal temperature in the two phases d = d = d ), we 
                                                                                    l           v
can write: 

         
       - d + v dp = 
                                                        - d + v dp  (5.15) 
             l                 l              l         v    v    v


As state equations for water vapour dry air and their mixture, i.e., moist air, 
one can use the state equation for ideal gas and Dalton's law as explained in 
Chapter 2. 
These relationships are, however, valid when the range of pressures and 
temperatures of the analysed process is close to the ambient ones. 

Derivation of the Kelvin equation 
The Kelvin's formula can be obtained remembering that the capillary 
pressure pc is defined as: 

        c
       p = v
                       p - p = 2H   (5.16) 
                                         l

in fact, in the following analysis the effect of meniscus curvature also 
appears through 2H, which is the same as the capillary pressure because 
we suppose that the mechanical equilibrium exists, i.e., the Young-Laplace 
equation (eq.(5.8)) is valid. 
In differential form: 

       dp = dp - dp = d (
             c            v                                  2H  ) (5.17) 
                                                   l

The molar based Gibbs-Duhem equation (5.13) for the liquid l and its vapour 
v phase becomes: 




 177


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 



        d - v dp + dµ = 0
         l                   l         l           l                                   (5.18) 
         v
        d - vv d v
                                       p + d v
                                                   µ = 0

and, since there is chemical equilibrium (5.11) i.e. dµ = dµ = dµ  and 
                                                                                                                                                        l    v
considering a constant temperature process (d = 0 ) eq. (5.18) becomes: 

       v dp = v dp                                                                     =                                            (5.19) 
        l          l              v         v                                                             constant

and substituting (5.17) in (5.19) to eliminate pl: 

                                                   v - v                                                                            v
       d (2H  )                                        l                        v                                                  v
                                  =          c
                                       dp =                                                     v
                                                                                      dp = d v
                                                                                                                         p -                  v
                                                                                                                                              dp  (5.20) 
                                                                  v                                                                 v
                                                                       l                                                                 l

where as we mentioned  is a weak function of H. By assuming a perfect gas 
behaviour for the vapour, i.e.: 

                        v
                         R                                           v
                                                                                                    v
                                                                                                p                        1
         v
       p =                                 or                                         =                                        (5.21) 
                        Mw                                        Mw                            R                       vv

we have: 

                                                                                      R
       d (2H  ) = d c
                                            p =                   v
                                                   dp -                                                        v
                                                                                                     dp ,         = constant (5.22) 
                                                                                 v p
                                                                                      l v

which can be integrated from pv=pv0 and pc=0 for H0=0 (i.e. flat surface) to pv 
and pc for H0. By assuming that v  remains constant, i.e. the liquid water 
                                                                                                                    l
compressibility tends to zero,    0  we have: 
                                                                                           l

                                                                                                     R                   p 
       (2H  ) = p = ( p - p ) -                                                                                    ln v 
                                       c           v                        v                                                      ,     and v  are kept constant
                                                                                                                                   
                                                                                 0                                                                 l
                                                                                                     v                    p 
                                                                                                          l                    v 
                                                                                                                               0
        (5.23) 

This is the Kelvin equation and shows that for the case considered here the 
vapour pressure decreases as H (curvature of meniscus) increases (or the 
mean radius decreases). Usually the first term on the right hand side of 
(5.23) is negligible. In fact, for water at 18°C and a mean radius of curvature 
of 1µm, we have pv/pv0=0.9891, and for a mean radius of curvature of 100, 
this becomes 0.897. 
If we allow for the change of v  with the pressure, a correction can be added 
                                                                                      l
to (5.23). However, this correction is generally small. To show this, we 
integrate (5.4) to arrive at: 

              v
       ln l = 
                             - (p - p ) (5.24) 
                                       l      l         l
              v                                              0
              lo




 178


                        Physics and thermodynamics of fluids in concrete at high temperature 




Now we note that for water   =5 10-11 dyne/cm2, and therefore the change 
                                                               l
in  v  is negligible. Other consequences of effect of curvature in a single 
      l
component system are the reduction of chemical potential, the increase of 
heat of evaporation and the change in triple point temperature. 
Clapeyron equation and Kelvin equation are strictly valid only for a pure 
vapour. For the case of a liquid in equilibrium with a mixture of ideal gases 
having a total pressure pg, to which is applicable Dalton's law 
 g         gw               ga
p = p                 + p , instead of equation (5.23) we obtain: 

                                                                                                      gw
                                                                         R                     p          
           ( H  )                     c
                            = p = ( g                   g 0
            2                               p - p ) -                              ln                       (5.25) 
                                                                              w                 gws
                                                                         v                      p          

For equilibrium between water and moist air, the total gas pressure can be 
usually considered constant (pg=pg0=atmospheric pressure). Moreover, in 
ordinary situations (pgªatmospheric pressure, i.e. about 1 bar) since the 
molar volume of the liquid phase  w
                                                                                   v  is much smaller than the specific 
volume of its vapour  gw
                                            v     , the influence of the total gas pressure pg on the 
vapour tension of the liquid pgws can be neglected. In such a case, equation 
(5.23) can be used, but now pv has the meaning of the partial vapour pgw 
pressure in the mixture and pv0 of the vapour tension of the liquid pgws (pgws 
= pgw for a flat surface, when pc=0). These hypothesis are surely valid in the 
normal range of variation of pressure and temperature, i.e. far away the 
critical point of water. But analysing the behaviour of concrete at high 
temperature it is clear that the situation is different. When temperature 
reaches values close to the critical point of water, an additional increase of 
water degree of saturation, caused by considerable thermal dilatation of the 
liquid water, is observed. The last phenomena cause an effect, referred 
sometimes as "moisture clog", and are a reason of existence of the four 
zones, where moisture content is higher than the initial one, and where 
even fully saturated conditions may be sometimes encountered. In 
particular the water vapour does not behave as a ideal gas in the range of 
temperature between 160°C and the critical point of water. If we consider 
the vapour as a perfect gas for calculating its density from the saturated 
pressure, i.e., in flat conditions as far as the interface is concerned, we 
obtain values two-three times lower than real values available from 
thermodynamic tables, see Figure 5.6-5.7. Considering the correct variation 
of molar volume or density with temperature, e.g. to define  gw
                                                                                                                                    v     or gw as 
mainly function of temperature, one obtains: 

                gw           gw                     R                                  
           v          = v ( , gw
                                        p )                                       gw
                                                 =                 + v
                                                                                                           for                <  < 
                                                           gw                                                             ref            crit
                                                    p                                                                                            (5.26) 
                gw               gw                                M
                                                                                                     
                       =               ( , gw
                                         p )              gw            w                      gw
                                                 =  p                        + 
                                                                                                                    for       <  < 
                                                                                                                          ref               crit
                                                                   R                                 





 179


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




Where  qref is the reference temperature (around 160°C) the additive terms 
          gw
     v
                and                                                               gw
                                                                              
                                                                                         are exclusively function of temperature taking into account 
the real behaviour of water vapour in the range 160-373.15°C (Figure 5.6-
5.7). The firsts terms on the right hand side of eq. (5.26) are the terms 
related to ideal behaviour. 
 

                                                                                    10


                                                                                         1


                                                /mol]                              0.1
                   3                                                                                                          Water vapour
                                                                                                                              Liquid water
                                                                                 0.01


                                                                              0.001

                                                         Molar volume [m

                                                                             0.0001


                                                                            0.00001
                                                                                    273.15 323.15 373.15 423.15 473.15 523.15 573.15 623.15 673.15

                                                                                                               Temperature [K]



                  Figure 5.6 Water vapour and liquid water molar volume: experimental 
                  data 



                                                                            4.E+02


                                                                            3.E+02

                        ]3 3.E+02                                                              Exper.values
                                                                                               Ideal Gas
                                                                            2.E+02             Numerical approx.


                                                                            2.E+02


                                                                            1.E+02
                        Vapour density [kg/m

                                                                            5.E+01


                                                                            0.E+00
                                                                                         0    50     100       150    200    250    300       350    400

                                                                                                               Temperature [°C]

                  Figure 5.7 Comparison of the experimental values for water vapour 
                  density with those calculated by using (5.26) and by using ideal gas 
                  hypothesis 



 180


             Physics and thermodynamics of fluids in concrete at high temperature 




Energetic meaning of capillary pressure 
Above critical point of water, the liquid water and water vapour do not exist 
and there is only the gas phase. As a result, there are no phase changes of 
the pore water (condensation ­ evaporation) and capillary pressure has not 
any physical meaning. Thus it cannot be directly used as a state variable for 
water, as for example vapour pressure does. Hence in these conditions we 
deal with a kind of the Stefan problem, where two regions of concrete, with 
the temperature below and above the critical point of water, are separated 
by the moving boundary. In these two regions the hygro-thermal state of the 
material is described by the different sets of state variables and equations 
expressing conservation laws of mass and energy for the system.  
There are some numerical methods for dealing with the Stefan problem, 
e.g., but they are numerically costly in the case of 2-D, and especially, 3-D 
problems. In this chapter a method which formally allows us to avoid direct 
tracing of the boundary position in the space will be presented. We give 
different physical meaning to the capillary pressure in the zone, where 
temperature exceeds the critical point of water, and apply a special 
`switching' procedure of the finite element from the below- to the above 
critical temperature description of the medium. 
 Above the critical point of water the moisture may be present in the pores of 
concrete in two forms, as the adsorbed, liquid water and the gaseous water. 
The latter forms a mixture with the dry air and is assumed to behave as an 
ideal gas. We assume that the adsorbed water is in thermodynamic 
equilibrium with the gaseous phase of water, thus it may be described by 
the equation [21],[22]: 

                          gw
                    p 
        = R ln                     (5.27) 
                          gws 
                    p          

where  = - Hws / Mw is the so called water potential, Hws ­ the enthalpy 
of water adsorption by the concrete skeleton, pgw ­ partial vapour pressure 
and pgws ­ equilibrium vapour pressure at a flat adsorbate surface). Below 
the critical point of water, Tcr, the value of the water potential is related to 
capillary pressure according to: 

      pc  = -   w (5.28) 

where w is the liquid water density. 
The surface diffusion of the adsorbed water on the solid skeleton may be 
described by: 

       j = -D grad   (5.29) 
       w      b


with jw being mass flux of the adsorbed water. 




 181


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




For the gaseous water there are two transport mechanisms: advection and 
diffusion, described by a similar formula as the one for the situation below 
the critical temperature of water. 
Using equation (5.27), one may substitute in previous equations vapour 
pressure and its gradient with the appropriate expressions containing water 
potential . 
The latter is in fact the state variable which describes thermodynamic state 
of both the adsorbed water and the water vapour. 
Because of lack of any trustable data concerning the equilibrium vapour 
pressure pgws  above temperature Tcr, we assume that it follows the perfect 
gas law, i.e. increases linearly with temperature. For the same reason we 
assume that the density of adsorbed water above the critical temperature 
Tcr is equal to the water vapour density, which is equal to the water density 
at the critical point (Fig.5.5-5.6). 
Now, using (5.28), we introduce formally into equations describing mass and 
energy transport the capillary pressure to obtain the same form of these 
equations as below the temperature Tcr. One must however remember that 
the capillary pressure only substitutes formally the water potential  and 
they have different physical interpretation.  
Because of different physical meaning of the capillary pressure pc  and 
different nature of the physical phenomena above the temperature Tcr (there 
is not liquid water ­ vapour phase change and associated with it heat 
effect), we have introduced a special `switching' procedure, which is applied 
to the whole finite elements. When part of FE nodes have temperature 
above Tcr, the capillary pressure is blocked at the previous value, until the 
temperature in all the nodes does pass the critical point of water. 

5.2.4  Enthalpy of vaporization 
The enthalpy of evaporation depends upon the temperature and may be 
approximated by the Watson formula, [13],[14]: 

                             5
        va
          h p = 2.672  10 ( -  )0.38
                                             cr      (5.30) 

where qcr= 647.3 K is the critical temperature of water. A comparison of the 
experimental data from [23],[24] and those calculated by using equation 
(5.30) is shown in Figure 5.8. 

5.2.5  Viscosity of moist air 
The dynamic viscosity of moist air, which depends upon the temperature 
and the ratio of the vapour and gas pressures, using the data from [23],[25], 
can be approximated using the following formula [26]: 

                                        ga 0.608
         g       gw                p 
       µ = µ +( ga
                       µ - gw
                           µ )              
                                                   (5.31) 
                                        g
                                   p 




 182


                                                 Physics and thermodynamics of fluids in concrete at high temperature 



                                                             gw
where                                                   p
            (1 - x ) =                                              is the mole fraction of dry air in the gas phase, 
                                                             g
                                                        p
     gw                                  gw 0
µ          = µ                                   + v
                                                    (T -T ) is the water vapour dynamic viscosity,  
                                                                     0
with µgw0= 8.8510-6 Pas, v=3.5310-8 PasK-1,  
     ga           ga 0
µ          = µ                                   + a
                                                   (T -T ) + a
                                                                              (T -T )2  is the dry air dynamic viscosity,  
                                                                     0                       0
with µga0= 17.1710-6 Pas, a= 4.7310-8 PasK-1, a= 2.2210-11 PasK-2.  
A comparison of the experimental data of Mason and Monchick [45] for 
moist air with those calculated using equation (5.31) is shown in Figure 5.9. 

5.2.6  Viscosity of liquid water 
The dynamic viscosity of liquid water depends strongly upon temperature 
[13] and can be evaluated, with sufficient accuracy in a wide temperature 
range, using the following approximate formula, [27] 

                  w
             µ =                                                   (T -           ) 1.
                                                                                   - 562
                                                  0.6612                    229               (5.32) 

which results are compared with the experimental data [13],[24] in Figure 
5.10. 
 

                                          3.00E+06

                                                                                                    Watson formula
                                          2.50E+06                                                  Exper. values


                                          2.00E+06


                                          1.50E+06


                                          1.00E+06
                  Latent heat  [J/kg]


                                          5.00E+05


                                          0.00E+00
                                                        273         323      373      423    473    523    573    623    673

                                                                                     Temperature  [K]



             Figure 5.8 Comparison of the experimental values for evaporation 
             enthalpy with those calculated by using the Watson formula (5.30) 





 183


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 





                                                                                                30                                                                dry air
                                                                                                                                                                  x=0.2
                                                                                                                                                                  x=0.4
                                          ]                                                                                                                       x=0.6
                                                s 25
                                               Pa                                                                                                                 x=0.8
                                                                                        µµµµ
                                                     ty [                                                                                                              pure 
                                                                                                20                                                                vapour

                                                             scosi
                                                                                                15
                                                                      ic vi


                                                                                                10
                                                                               Dynam


                                                                                                 5
                                                                                                      0         50      100       150     200        250        300

                                                                                                                         Temperature  [oC]

           Figure 5.9 Comparison of the experimental data for the dynamic 
           viscosity of moist air with those calculated using approximated formula 
           (5.31) 



 
 

                                                                                          0.0018
                                                                                          0.0016                                                      Exper. values
                                                                                                                                                      Thomas
                                                                                          0.0014
                a s]
                                                                                          0.0012
                        ty [P
                                                                                                0.001
                                 scosi 0.0008
                                 ic vi 0.0006
                                                                                          0.0004
                                 Dynam
                                                                                          0.0002

                                                                                                      0
                                                                                                      273.15          373.15        473.15          573.15        673.15

                                                                                                                                Temperature  [K]



           Figure 5.10 Comparison of the experimental data for the dynamic 
           viscosity of water with those calculated with Thomas formula (5.32) 





 184


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 





5.3  Concrete at high temperature: heat and mass 
        transfer, inner structure and thermal 
        properties 
When hygro-thermic and mechanical behaviour of concrete at high 
temperature, for instance during fire, is analysed, several complex and 
interrelated physical phenomena should be taken into account. Phase 
changes and dehydration play an important role in the mass and heat 
transfer during these processes [28]-[31] and [32],[33]. 
In a concrete element exposed to fire conditions a moving front, separating 
four well defined zones, may be observed. The first one, close to the heated 
surfaces, is almost completely dry and has an increasing temperature. In 
the second one, in the internal part of the heated element, both moisture 
content and temperature are almost unchanged and their values correspond 
to the initial state of the material. Between these zones, there is area with 
high gradients of temperature and moisture content, causing considerable 
thermo-diffusion flux of water, which results in an increased moisture 
content inside a concrete element, close to the moving front, Figure 5.11. 
 
                                                                                                           drying/dehydrating
                                                                          dry/dehydrated                              quasi-saturated initial state


        x                                                                     x
             lu                                   concrete                                                                       concrete
                                                                                   lu
                    f                              element                                f                                      element
                         eat                                                                   eat
                                h                                                                     h





                                                                                                                          pressure

                                           temperature

                                                                                                                                   temperature
                                           pressure

                                         distance from heated face                                              distance from heated face               

                                     Figure 5.11  The four observed zones and the process for the build-up of 
                                     pressure 



Usually, slow heating causes water migration to occur in the both the liquid 
and vapour phases, and generally leads to zones of physical saturation as 
drying progresses. The zone of saturation itself migrates under the pressure 




 185


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




difference corresponding to the saturation vapour pressure at its two faces, 
and increases in length as condensation occurs at the hottest face. 
Rapid heating leads to early physical saturation and a rapid build up of 
water at the hottest face of the physically saturated (or quasi-saturated) 
zone. This is frequently coupled with further increases of temperature which 
causes expansion of the water in the saturated pores, due mainly to the 
increase of its density close to critical point (see 5.2.1 and 5.2.3), and 
fracture of concrete. 
In previously unheated concrete the application of non-uniform 
temperatures elevates the pore vapour pressures and causes initial 
migration towards lower temperatures zones. Initially, for concrete in as 
cast conditions, the internal pores are not physically saturated. Typically 
they contain an unfilled pore volume of 2-4% for normal concrete. The 
corresponding figure for high performance concrete is likely to be much less, 
particularly when superplasticisers are used to produce a flowing mix. 
Theses unfilled pores play an important role in the early migrational 
behaviour by acting as reservoirs or sinks to moisture movement and 
preventing the build up of excessive pressures during the temperature 
transient phase of heating. 
After the early moisture migration has filled some of the initially unfilled 
pore volume, a "saturation plug" (also known as "moisture clog") forms and 
water then moves under the action of a hydraulic gradient governed by 
saturation vapour pressures at the two ends of the plug. The pressures in 
this zone are generally greater than the pore pressures before saturation 
occurs. Under steady-state conditions the saturation zone moves away from 
the hottest region and increases in length simultaneously as more water 
accumulates at the upstream face. 
The phenomenon of moisture clog may be a cause of possible explosive form 
of spalling of the concrete, because of the high values of water vapour 
pressure reached during the heating process, as it will be explained in the 
sections concerning the mechanical behaviour of concrete in high 
temperatures conditions. Anyway, if concrete is not in as cast conditions, i.e. 
the initial water content and saturation degree are much lower, the risk of 
the saturation plug occurrence is considerably reduced and, consequently 
the explosive spalling hazard is reduced too. 
In the above described processes, the movement of water inside the material 
involves several types of physical mass transport in the pores of heated 
concrete. 
In the following sections an analysis of these different transfer modes and 
the related non-linear coefficients will be done. 

5.3.1  The Darcean flows 
For fluid phases the multiphase Darcy's law has been applied as 
constitutive equation (see 2.4), thus for capillary water we have: 





 186


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 



                                w
               w           k
          v = -                      (              g            c              w
                                               p
                                                        - p
                                                                     -              b)       wg g      w
                                                                                          + k v + k           
                                                                                                                 (5.33) 
                                w
                           µ

while for the gas phase holds: 

                                g
               g           k
          v                          (         g                 c              g
                                          p                p                         b)      gw w       g 
                                                                                                       
                    = -                                 -           -                    + k v + k           
                                                                                                                 (5.34) 
                                g
                           µ

where kg and kw are permeability tensors of the gaseous and liquid phases, 
vw and vg are the volume averaged velocities of capillary water and gaseous 
phase relative to the solid phase and the vector b indicates the specific body 
force term (normally corresponding to the acceleration of gravity). 
The second terms in the right hand side of (5.33) and (5.34) represent the 
viscous drag of one fluid upon the other and kwg and kgw are the related 
permeability tensors. Therefore it is expected that kwg and kgw depend of the 
viscosity ratio  w / g
                                     µ              µ , among other parameters. For example, when 
 w / g
µ     µ   , the viscous shear exerted on the w-phase by the g-phase will 
be negligible. Considering water as liquid phase and moist air as gasous 
phase we have  w / g
                                     µ         µ                1 , so these terms related to viscous drag between 
two phases can be omitted. 
The third terms in the right hand side of equations (5.33) and (5.34) 
represent the effect of surface tension gradient. In fact, the tangential force 
balance shows that the gradient of the surface tension can influence the 
interfacial motion. The result is an extra driving force in the volume 
averaged momentum equations (Marangoni effect). The coefficient of this 
term, a new conductance, is related to the matrix and fluid properties. 
However, the evaluation of these additional terms, especially in high 
temperature conditions, appears to be really difficult, so also these terms 
will be neglected in the formulation of Darcy's formula. 
As far as the remaining coefficients in equations (5.33) and (5.34) is 
concerned, kw and kg are the most studied coefficients and almost all studies 
are empirical. 

Relative permeability 
Muskat and Meres (1936) recommended that the phase permeabilities kw 
and kg be treated as isotropic and given by: 

               w           rw
          k = kk  (5.35) 
               g           rg
          k = kk

where k is the absolute (or intrinsic) permeability used in the single phase 
flows and krw and krg are the relative permeability of the w and g phases, 
respectively. 
Then the relative permeability accounts for the phase distributions, i.e.,: 




 187


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




            r           r                                         
       k          = k  matrix structure, S,  ,
                                                 ,
                                                        ,  history
                                                                           
                                                            
                                                                   

where  S is the liquid phase saturation,  the surface tension and J is the 
contact angle. 
Therefore, the approach taken by Muskat and Meres was the simplification 
of this relation to a saturation dependence only, i.e.,: 

            rw           rw
       k          = k (S ) (5.36) 
            rg           rg
       k          = k (S )

which is still in popular usage. 
However, these coefficient are defined for the Darcy (Stokes) flow with a 
negligible interfacial drag at Awg, and a negligible Marangoni effect 
(  0 ). When necessary, in some studies, these effects are included 
directly in the formulation of relative permeabilities. 
Since the simultaneous flow of the two phases has been made analogous to 
that of the Darcean single-phase flow, the flow of each phase at given local 
saturation depends on how the two phases are distributed over the 
representative elementary volume (REV). The factors that can be 
influencing the phase distributions are the local saturation, matrix (or pore) 
structure, history, contact angle, surface tension and density ratio. These 
effects are generally determined experimentally. 
The local saturation, i.e., the extent to which the liquid fills the pores in the 
REV, influences the phase distributions in this volume and in turn 
influences the resistance to the flow of each phase. Considering the 
complexity of the phase distribution function, the local flow channels 
through which each phase flows is a complex function of S. 
The  excess resistance, i.e, 1/krw, also becomes a complex function of local 
saturation. One of the important features of the excess resistances 1/krw and 
1/krg is that the simultaneous flow of the phases causes excess resistances 
for both phases. This means that in general for a given saturation we have: 

            rw           rg
       k          + k          < 1  (5.37) 

The intermediate saturations correspond to a mutual and significant 
resistance offered by each phase to the flow of the other phase. 
The solid phase distribution influences the distribution of the w and g 
phases and, in turn the relatives permeabilities. Since the w and g phases 
flow through the passages that are nonuniform in cross section, the 
distributions of the phases (and their point velocities) are significantly 
influenced by the topology of the passages. The absolute permeability K is a 
measure of the average throat size, but it does not give all the information 
about the microstructure. 





 188


    Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




As far as the history is concerned, phase distributions experience hysteresis 
depending on whether a given saturation is reached trough a drainage or 
imbibition branch. 
As expected, also wettability influences both the dynamic and quasi-static 
phase distributions, while when the pore length scale is small, the surface 
tension can significantly influence the meniscus contur. 
The density ratio accounts for the effects of the buoyancy on the phase 
distributions. 
Anyway, the simplest way to evaluate relative permeabilities coefficients, is 
to consider a simple dependence on the only saturation, similar to equation 
(5.36). 
In this sense, two different approaches exists, depending on a pendular 
region existing, or not (region Ib in Figure 5.1). By taking into account an 
irreducible saturation level Sirr, the liquid phase continuity is disrupted 
when the saturation falls below Sirr, so that liquid flow is not longer 
possible. This behaviour is described by introducing a krw model which can 
be represented as follows: 

                                         Aw
                    S - S 
            rw                    irr
       k          =                     
                                             for S>S
                                                                 (5.38) 
                     1                                  irr
                           - S           
                                  irr 

where  Sirr is the irreducible saturation, which is discussed in detail in 
[12],[34] and Aw is a constant, with the value from the range <1,3>, 
[2],[12],[18],[34]. 
On the other hand, if Sirr is zero, i.e., the liquid phase remains continuous, 
i.e., in funicolar state, during the overall nonhygroscopic region, one has the 
following form of relative permeability for the liquid phase, see Figure 5.12: 

            rw           Aw
       k          = S           (5.39) 

Where, typically the exponent Aw varies in the range <1,6>. 
The first formulation. i.e., with irreducible saturation, there is not liquid 
phase transport in the pendular zone. This means that in this region there 
is only vapour diffusion flux driven by vapour concentration. Withaker has 
shown that in such condition the convective flux, driven by pressures, is not 
enough to overcome the diffusive flux, which is from the region of high 
temperature to region of low temperature. But experimentally it is known 
that drying processes globally is a loss of mass from the inner part of 
material toward the exposed surface. Because of this in the mathematical 
model a singularity is produced. 
In the second formulation, i.e., without the irreducible saturation concept, a 
continuous moisture flux toward the surface exists, also between Sirr and 0 
(solid saturation point) and in such case the problem of singularity is 
avoided. In reality, a sharp increase of the capillary water flux is observed 
for concrete, when the relative humidity RH reaches the value higher than 
about 75% [35]. 



 189


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 





                                                      1

                                                   0.9

                                                   0.8

                                                   0.7

                                                   0.6
                                                                       Water
                                                   0.5                 Gas
                                                   0.4                 Sum

                                                   0.3

                                                   0.2
                  Relative permeabilities [-]
                                                   0.1

                                                      0
                                                             0         0.2                    0.4        0.6     0.8         1

                                                                                              Saturation [-]



            Figure 5.12  relative permeability for liquid water, gas and their sum, 
            according to formula (5.39) with Aw=6 and (5.41) with Ag=1, Scrit=1 



Such a kind of behaviour can be described by the following relationship, 

                                                                 -           
            rw =  +                                                                           (5.40) 
                                                            (1 RH
       k                                         1                                       w
                                                                                         A
                                                                               S
                                                                  0.25 )           -1
                                                                        Bw

                                                                              
                                                                             

where Bw is a constant, see Figure 5.13. 


                                                      1
                                                   0.9
                                                   0.8
                                                   0.7
                                                   0.6
                                                  [-] 0.5
                 rw k 0.4
                                                   0.3
                                                   0.2
                                                   0.1
                                                      0
                                                             0         0.2               0.4           0.6      0.8     1

                                                                                    Relative humidity [-]



            Figure 5.13 relative permeability of liquid water as function of R.H. 
            according to equation (5.40) 



 190


    Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 





                                 1

                              0.9

                              0.8

                              0.7
                              0.6                              298.15
                                                               333.15
                             [-] 0.5
                    rg k                                       373.15
                              0.4                              413.15
                                                               453.15
                              0.3                              513.15
                              0.2                              633.15

                              0.1

                                 0
                                        0                      0.2       0.4          0.6    0.8    1

                                                                         Saturation [-]



              Figure 5.14 relative permeability of gas according to equation (5.41) 
              with Scrit=Scrit(T) and Ag=3 



The term in square brackets has a form originally proposed in [35] for the 
water diffusivity (with Bw=6 or 16). It has good numerical properties and 
allows to avoid a use of irreducible saturation concept which creates serious 
theoretical and numerical problems [12]. 
The gas relative permeability for concrete, similarly as for most capillary 
porous materials, may be described by the formula of the form, 
[2],[12],[18],[27]: 

                                                         Ag
              rg                              S 
         k            = 1 -                             
                                                        ,                     for S< cr
                                                                                     S  (5.41) 
                                                        
                                                   cr
                                                   S 

where Scr, variable with temperature, is the critical saturation value, above 
which there is no gas flow in the medium, and Ag is a constant, which 
usually has value from the range <1,3>, [2],[12],[18],[27], see Fig. 5.14 and 
5.15 

5.3.2  Adsorption-desorption phenomena: isotherms of desorption 
For a given matrix structure the amount of the gas adsorbed on the internal 
surface of the matrix is proportional to the volume of the matrix as long as 
the sampling volume is large (many times the representative elementary 
volume). If the density does not change noticeably, then we can express the 
adsorption capacity of the matrix by: 




 191


                                             Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 



                                       1                                                                                             1

                                      0.9                                                                                           0.9
                                                                                 Gas                                                                                         Gas
                                      0.8                                                                                           0.8
                                                                                 Water                                                                                       Water
                                      0.7                                                                                           0.7
                                                                                 Sum                                                                                         Sum
                                      0.6                                                                                           0.6
                     [-] 0.5                                                                                    [-]
                                                 T=25°C                                                                            0.5         T=100°C
     r                                                                                         
                    k                                                                               r k
                                      0.4                                                                                           0.4

                                      0.3                                                                                           0.3

                                      0.2                                                                                           0.2

                                      0.1                                                                                           0.1

                                       0                                                                                             0
                                             0      0.2         0.4       0.6     0.8     1                                                0      0.2      0.4       0.6      0.8     1
                                                               Saturation [-]                                                                              Saturation [-]



                                       1                                                                                              1
                                      0.9                                        Gas                                                0.9                                      Gas
                                      0.8                                        Water                                              0.8                                      Water
                                                                                 Sum
                                      0.7                                                                                           0.7                                      Sum

                                      0.6                                                                                           0.6
                          [-]                                                                                           [-]
                                      0.5         T=200°C                                                                           0.5
                                                                                                                                              T=300°C
               r                                                                                          r
                                 k                                                                                             k
                                      0.4                                                                                           0.4

                                      0.3                                                                                           0.3

                                      0.2                                                                                           0.2

                                      0.1                                                                                           0.1

                                       0                                                                                              0
                                             0      0.2         0.4       0.6     0.8     1                                                0      0.2      0.4       0.6      0.8     1
                                                               Saturation [-]                                                                              Saturation [-]



                                                   Figure 5.15  relative permeability of water, gas and their sum according 
                                                   to equation (5.39) with Aw=6 and (5.41) with Scrit=Scrit(T) and Ag=2, for 
                                                   several temperature: 25, 100, 200, and 300 °C 



                                                  g  of gas adsorbed                *
                                                                                  m  (5.42) 
                                                           g  of solid              ad


     where m*ad is also called adsorption isotherm and is depending on pg, i.e. the 
          pressure of the gas, temperature T, molecular properties of the gas and 
          solid, matrix structure, history. Experimental results show that m*ad is not 
     necessarily single valued and after the monolayers are formed and if the 
     pore size is in a specific range, then it can be history dependent due to the 
     capillary condensation. In general, m*ad(T,pg) < m*des(T,pg) for the same 
     value of pg and T, i.e., the amount of mass adsorbed is larger in the 
     desorption branch. 
     Typical forms of isotherms for concrete at ambient temperature are reported 
     in Figure 5.16. The lack of the inflection in Fig. 5.16b at low pressures, is 
     due to the building of the higher layers before the completion of the 
     monolayer. 
     A schematic classification of the pore size in porous material could be the 
     following: 



           192


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




Micropores  with d < 20 Å 
Mesopores  with 20 < d< 500 Å 
Macropores  with d > 500 Å 
Note that in the practice a range of pore sizes is present, i.e., the micro and 
meso pores coexist, hence, the adsorption isotherms demonstrate an average 
behaviour. 
 





          *                                                         *
                    ad                                                        ad
               m                                                         m



                                       inflection 


                                          p
                            0 1                                                     0               1
                                            g/pg0                                        pg/pg0 

                                           a)                                              b) 


                          Figure 5.16 possible types of isotherm with hysteresis: a) with 
                          inflection and b) without inflection 



When the Kelvin relation for the vapour pressure lowering (due to meniscus 
curvature) is valid in the range of pore sizes, the adsorption isotherm is 
used for the pore size prediction. About this possibility, some observation 
are necessary looking at Figure 5.17: 
     ·  there is a transitional relative pressure (pg/pg0)t where above this 
        transitional value the adsorption by a mesoporous solid becomes 
        distinct from that for the nonporous solid. This transition is shown in 
        Figure 5.17 ; 
     ·  for (pg/pg0)< (pg/pg0)t the adsorption is restricted to a thin layer on 
        the solid surfaces. Therefore, no distinction appears in the general 
        behaviour of m*ad in this regime; 
     ·  for (pg/pg0)< (pg/pg0)t the capillary condensation begins starting from 
        the smallest pores (as suggested by the Kelvin equation); 
     ·  when (pg/pg0)=1, almost all the pores are filled with the liquid phase, 
        and the saturation tends to unity; 
     ·  the hysteresis loop in some system can appear with point F moving 
        toward the right; 
     ·  for pore sizes (and meniscus curvatures) less then 20 Å the Kelvin 
        equation is not expected to hold, because the interfacial layer 
        thickness in not substantially smaller than the pore size. 


 193


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




 



                                                                                                                1 
                                                                                                          F     G 
                                                                                      E' 





                                                                                                E 
                                                          Evaporation 
                * ad                                                                           Condensation           S 
                          m



                                                                      D 
                                                          C 
                                             B 
                                A             Multilayer                         Capillary 
                                              adsorption                         Condensation -evaporation      0 

                                                                            p
                                        0 1 
                                                                            g/pg0 

                                                   (pg/pg0)t 


                                      Figure 5.17  typical isotherm curve with hysteresis loop 



The increase of (pg/pg0) with increase in m*ad can be tentatively explained by 
the Kelvin relation. This states that as the condensation proceeds and the 
film thickness and radius of curvature of the liquid pore increase, the 
vapour pressure approaches the saturation pressure. 
Also the observed (pg/pg0)ad >(pg/pg0)des for a given m*ad and T is generally 
explained in terms of Kelvin equation and by using the observation that 
condensation begins at the solid surface after a thin adsorbed film is created 
which is the nucleation surface of the condensation. In order to use the 
Kelvin equation, geometrically specific arguments are used in which it is 
shown that the curvature during the adsorption and desorption branches is 
different. The simple form of Kelvin equation can be written as. 

          p                             2 w
               g                                     v
                                                                
                               = exp-                    H 
                                                                (5.43) 
                                                                
           g
          p                                        R
                0


where 





 194


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




                        1  1                   1 
        H                                            
                   =                      +            (5.44) 
                        2                            
                          c
                               R                 c
                                                R 
                                    1            2


As we have briefly mentioned in 5.2.3 the interface curvature can be for 
spherical meniscus (capillary water) equals to 

         1              2  cos 
                   =                             (5.45) 
          c
         R               r - a
                                      t

i.e. Rc1=Rc2, while for cylindrical interface (bound water) it is  

          1               1
                   =                        (5.46) 
         R              r - t
              c                  a


i.e., one of the two curvature radius is infinite. 
 




                                                           a





                                                           b





                                                           c





                                                           d


          Figure 5.18  scheme of capillary condensation process in a cylindrical 
          tube 



We need to show that Had<Hdes. To proof this, we can consider the capillary 
condensation process inside a cylinder open on both sides Fig. 5.18. 




 195


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




In a) the pore is in equilibrium with vapour at low (pg/pg0), curve AB in Fig. 
5.17, and there are adsorbed molecules on the pore walls. In b) a situation is 
shown where the adsorbed layer has become liquid-like. However, by 
Kelvin's relation, eq. (5.43), if an attempt is made to keep (pg/pg0) constant, 
the system is unstable since the mean curvature of the cylindrical interface 
(eq. (5.46)) within the pore increases as the amount adsorbed increases, i.e., 
the pore will spontaneous fill as depicted by the curve DEF in Fig. 5.17. 
Filling will continue from the onset of instability until a situation as in c) is 
reached where fluid with spherically shaped interfaces is in stable 
equilibrium with the vapour phase. On reducing (pg/pg0), desorption will 
become spontaneous as soon as the radius of the interface equals that of the 
pore (curve FE'D in Fig 5.17). 
If we assume that (1/Rc1)ad<(2/Rc1)des then we have Had<Hdes. However, this 
assumption is not rigorously justified 
If we schematise a pore as a cylindrical tube, the situation during the mass 
movement inside the porous material is illustrated in Figure 5.19: 
 
                Bound water 


                             2
      3 1 
                                                                                                Vapour 
                                                                                                Fick diffusion 
                                                                                                Knudsen effect
           t-layer 
                                  R                                                             R
                                              Capillary condensate (liquid) 
             Vapour                                    Darcean flow 
          Fick diffusion 
         Knudsen effect 
                                       t2                                                      t1 
        Downstream                                                              Upstream 
         desorption                                                             adsorption 



           Figure 5.19  schematic representation of a capillary pore as a cylindrical 
           tube 



The degree of saturation with liquid water  S (considering together 
hygroscopic and capillary water, if the latter is contained in the pores), is an 
experimentally determined, function of capillary pressure (matric potential) 
pc and temperature T. Because of the difficulties to obtain from 
experimental tests such a kind of curve in temperature, particularly above 
100°C, and the processes which are of interest for our aims are heating of 
concrete, i.e. globally are drainage processes even if a local condensation is 
possible, hysteresis phenomena are not considered here, so isotherms  
curves are assumed to be a unique function of capillary pressure: 

         S = S(pc, T) (5.47) 

Usually the isotherms curves from water vapour desorption tests are in 
terms of content of water versus relative humidity. 



 196


   Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




The liquid saturation can be calculated from the mass water content w of 
the material, by using the following relation: 

                       s
                 w
          S =                (5.48) 
                       w
                 n

where s is the apparent density of the material and n is the total porosity. 
Moreover, the Kelvin's equation (5.7) can be used to have a relation between 
pc and the relative humidity. In this way it is possible to represent 
graphically the sorption isotherms in the form (5.47). 
The relationship (5.47)) is clearly fundamental for realistic modelling of 
hygro-thermic behaviour of the specific concrete, thus it must be determined 
during sorption tests at several temperatures [30],[31]. An example of such 
dependence, obtained using the concrete data from Bazant [30],[31] is the 
following: 

               = ( c, ) = { ( c
          S      S p T        h p ,T )}1/m(T )  (5.49) 

where  h is the relative humidity and m is a coefficient function of T 
determined correcting experimentally the values obtained from the 
thermodynamic tables of water and water vapour. 

                                                                      2
                                T                          T + 10 
          m(T ) = 1.04 -                       ,      T  =          
                                                                       (5.50) 
                            22.34 + T                     T + 10 
                                                                 0

the final form (5.49) is calculated from the original formulation given by 
Bazant in terms of w/c,  where  c is the cement content, and h through 
equation (5.48). 
Graphically the relationship represented by equation (5.49) in the capillary 
pressure domain is visible in Figure 5.20. 
Using such a kind of isotherms in a multiphase model like the present one, 
a problem can arise when the temperature builds up to critical point of 
water. In fact, Bazant's model follows a typical phenomenological approach 
where the single phases in the pores of material are undistinguished and 
only a general "fluid phase" is present in the voids of concrete. In this 
manner it is possible do not consider the critical point of water, simply 
keeping a certain quantity of water also above Tcrit. It is evident, as Bazant 
pointed out in his numerous papers on this subject, that when the 
temperature is under the critical one the water content in these isotherms is 
to consider as liquid water, i.e. free water plus a small percentage for bound 
water, while when T overcomes the critical point, the residual water has to 
be considered as gas water or a monolayer of physically bound water. While, 
with this formulation a consistent lot of water is present also above critical 
temperature, representing the gas phase and the chemically bound water in 
this zone which are separately computed in the model presented in this 
work. 



 197


      Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 





                                               100%

                                                    90%                                                                          25 °C
                                                    80%                                                                          75 °C
                                                                                                                                 100 °C
                                                    70%                                                                          150 °C
          [%]                                       60%
                 n                                                                                                               200 °C
                      tio                           50%                                                                          250 °C
                             ra                                                                                                  300 °C
                                                    40%
                                   atu                                                                                           350 °C
                                          S         30%

                                                    20%

                                                    10%

                                                         0%
                                                         1.00E+05                     1.00E+06     1.00E+07    1.00E+08      1.00E+09      1.00E+10

                                                                                                  Capillary Pressure [Pa]



                                                         Figure 5.20  Desorption isotherms according to Bazant [30], eq. (5.49) 



In fact, in a micro-mechanical approach where the three phases are clearly 
distinct it is necessary that, thinking to isotherms as curves of "liquid 
water", the value of saturation tends to zero approaching the critical point of 
water and it is practically zero above this point. 
For this reason, another formulation for desorption isotherms due to 
Baroghel-Bouny [36] has been considered. In this formulation the following 
relationship between capillary pressure and saturation, for "ordinary" 
concrete and high performance concrete at ambient temperature, is defined: 

                                                                                                         1
                                                    c               c
                                               p               p                                    -
                                                                         (S )         a ( b
                                                                                         S-
                                                         =                       =             - 1)(1 b ) (5.51) 

where the coefficients a and b in eq. (5.51) are material dependent according 
to table 5.2 where B and C stand for concrete and cement paste respectively, 
and O and H stand for ordinary and high performance concrete. 
 
                                                                                                               Mix 
     Coefficients                                                                     BO BH CO CH 
       a 
        [MPa] 18.6237 46.9364 37.5479 96.2837 
         b 
                                               [-]  2.2748 2.0601 2.1684 1.9540 


                                                               Table 5.2  parameters for eq.(5.51) at ambient temperature 





 198


    Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




The formulation given with equation (5.51) has been changed to take into 
account high temperatures effects, keeping constant the parameter b, and 
modifying the parameter a which is now a function of temperature, in the 
following manner: 
if T<100°C 

       a=costant 

if T>100°C 

       a=Q0+Q2 

where Q2 is a constant, while Q0 is a polynomial or an exponential function 
of temperature, 

                 (                           )( 1/ )
                                        1                        b
       S         G                                -
            =               +                                          (5.52) 

where G is a function of temperature and capillary pressure: 

                                                                                        b
                                                                       E         
       G = G (                               c
                            T p )                                                      b 1
                                   ,                                         c
                                                       =  p -
                                                                                                  
                                                                       a         

and where the parameter E is a function of temperature which has different 
form depending on the value of temperature: 

                                                                 N
                 T                     -T 
                           crit                   0
       E =                                                                               for T<Tcrit
                  T                    -T 
                  crit                                                                                                                 
                 N                                                          N                                     
       E =                 E T + E -                                                  E (T                    - z )     for T>T
                              0                                       0                      0         crit                     crit
                      z                                                     z                                     

where T0 is the ambient temperature, N is a parameter greater than 1 and 
the parameter z governs the transition through the critical point of water. 
The residual water content above the critical temperature is related to the 
bound water strongly adsorbed (in a monolayer state) on the walls of the 
pores (see also 5.2.3). 
The resulting isotherms from the above mentioned formulation for an 
ordinary concrete are represented in the following figure, in the capillary 
pressure and temperature domain respectively: 





 199


  Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 





                         1                                                          20
                        0.9                                                         75
                                                                                    100
                        0.8                                                         150
                        0.7                                                         200
                        0.6                                                         250
                                                                                    300
                        0.5                                                         350
                        0.4
      Saturation [-] 0.3
                        0.2

                        0.1

                         0
                        1.00E+04 1.00E+05 1.00E+06 1.00E+07 1.00E+08 1.00E+09 1.00E+10

                                               Capillary pressure [Pa]

                                                      a) 


                          1

                        0.9

                        0.8                                                                1.00E+05
                        0.7                                                                1.00E+06
                                                                                           5.00E+06
                        0.6                                                                1.00E+07
                        0.5                                                                2.00E+07
                        0.4                                                                5.00E+07
      Saturation [-]                                                                       1.00E+08
                        0.3                                                                1.00E+09
                        0.2                                                                1.00E+10

                        0.1

                          0
                               0       100          200           300        400

                                              Temperature [°C]

                                                      b) 


                        Figure 5.21  isotherms of desorption according to formula (5.52) in the 
                        capillary pressure domain a) and temperature domain b) for an ordinary 
                        concrete 



Figure 5.22 represents the desorption curves for a High Performance 
Concrete. 



 200


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




 


                                        1                                                              20
                                                                                                       75
                0.9                                                                                    100
                0.8                                                                                    150
                0.7                                                                                    200
                                                                                                       250
                0.6                                                                                    300
                0.5                                                                                    350

                0.4
          Saturation [-] 0.3
                0.2

                0.1

                                        0
                1.00E+05                                 1.00E+06     1.00E+07         1.00E+08    1.00E+09    1.00E+10

                                                                     Capillary pressure [Pa]

                                                                         a) 


                                              1

                                             0.9

                                             0.8                                                                 1.00E+05
                                             0.7                                                                 1.00E+06
                                                                                                                 5.00E+06
                                             0.6                                                                 1.00E+07
                                             0.5                                                                 2.00E+07
                                             0.4                                                                 5.00E+07
                      aturation [-]S                                                                             1.00E+08
                                             0.3                                                                 1.00E+09
                                             0.2                                                                 1.00E+10

                                             0.1

                                              0
                                                    0         100               200         300         400

                                                                     Temperature [°C]

                                                                         b) 


          Figure 5.22  isotherms of desorption according to formula (5.52) in the 
          capillary pressure domain a) and temperature domain b) for HP concrete 





 201


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




5.3.3  The bound water flow 
For the bound water flow, the following generalised constitutive law is 
applied [17],[37]: 

            bw                     bw
         v              = -D             S
                                           (5.53) 
            d                      d           b


where  bw                     bw
            D           = D (S ) is the bound water diffusion tensor, and the degree of 
                 d            d           b
saturation with adsorbed water, Sb, is given by 

         S = S                           for S  S
            b                                                                      ssp  (5.54) 
         S = S                                                            S > S
            b            ssp                        for                            ssp

According to the adsorption theory, the latter depends upon the partial 
vapour pressure [20] (or the relative humidity of the moist air), and because 
of the Kelvin equation (2.8), also upon the matric potential pc (see the 
remarks after equation (2.9) about the notation used): 

         S = S ( gw
                              p ) = S ( c
                                                         p ) (5.55) 
            b            b                          b

Equations (5.53) - (5.55) show clearly, that the driving force for the bound 
water flow is, in fact, the matric potential. The relationship (5.55), results 
from the adsorption theory, e.g. [20], where the adsorbed water content or 
volume fraction were introduced as additional state variables, not related to 
the vapour pressure value, hence also with the matric potential. 
In order to obtain a formulation containing directly a variable state (pc) eq. 
(5.53): 

                                         S
            bw                     bw          b         c
         v              = -D                        grad p  (5.56) 
            d                      d           c
                                          p
                                         

so the diffusion tensor multiplied for the derivative of degree of saturation of 
bound water on capillary pressure is a kind of equivalent diffusion tensor. 
The bound water diffusion coefficient for a concrete was found in [17] to be 
give as (Figure 5.23, isotropic material): 

                                                                    
              bw              bw                              S 
         D              = D exp-2.08                                (5.57) 
              d               d                                     
                              O                         S           
                                                              ssp 0 

where  bw
            D  =1.57x10-11 m2/s and reference temperature T
              d                                                                           0=295 K. 
                   O


5.3.4  Diffusion of water vapour 
For the description of the diffusion process of the binary gas species mixture 
(dry air and water vapour) Fick's law is applied (see 2.4.8): 
 



 202


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 



                                                                                                              ga
                ga                                 MaMw ga                                           p 
           v                    = -                                           D                                        
                g                                                                        grad
                                                                  2            g                     g 
                                                             M                                                          
                                                                                                                       
                                                                  g                                       p                                    (5.58) 
                                                                                                               gw
                                                    MaMw gw                                          p                                gw
                               =                                              D grad                                     
                                                                                                                         = -v
                                                                  2                g                               g                    g
                                                             M                                                            
                                                                                                                         
                                                                  g                                           p

                                                   gw                                   ga
                                                                                  
where  1                                                          1                            1
                                   =                                     +                                
                                                     g                                   g
           Mg                                    Mw                                     Ma
and  ga                                     gw
        g
      D                   = g
                                     D                  = D  is the effective diffusion coefficient of vapour species 
                                                                       eff
in the porous medium (isotropic material). 
 
                              1.6E-11

                ] 1.4E-11                                                                                                                     T=300 K
                /s2                                                                                                                           T=400 K
                              1.2E-11                                                                                                         T=500 K
                                                                                                                                              T= 600 K
                              1.0E-11

                              8.0E-12

                              6.0E-12

                              4.0E-12

                       Bound water diffusivity [m 2.0E-12

                            0.0E+00
                                                             0                          0.1                   0.2                      0.3    0.4         0.5

                                                                                                              Saturation  [-]


             Figure 5.23 dependence of the bound water diffusion coefficient for 
             concrete upon temperature and saturation according to equation (5.57) 



The effective diffusion coefficient of vapour inside the pores of partially 
saturated concrete varies with the saturation, temperature and gas pressure 
changes and may be described as, [17,21,39,42-44]: 

                                                                                                                             v
                                                                                                                        B
                                                                                                       p
           D                       = n (1 - S ) v
                                                                              A                                                   o
                                                                                        f D                                           (5.59) 
                eff                                                                      S     vo                             g
                                                                                                     o
                                                                                                            p

where Dvo=2.5810-5 m2/s is diffusion coefficient of vapour species in the air 
at the reference temperature o=273.15 K and pressure po=101325 Pa [17], 
Av is a constant, usually from the range <1,3>, [14],[15],[38]-[40], Bv is a 
constant, which for the value Bv=1.667 gives good correlation with the 
experimental data concerning vapour diffusion at different temperatures 
[25], and fS is the structure coefficient, which is introduced in the present 
model instead of previously used tortuosity factor, e.g. [34],[38],[41], and 



 203


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




will be discussed in next section. The temperature dependence of the 
effective vapour diffusion coefficient, evaluated according to equations (5.59) 
and (5.64), for the B-35 basalt concrete [42] (with  Av=1 and Bv=1.667) is 
shown in Figure 5.24. 


               ]  9.E-07
               /s2 8.E-07                                                                                                T=20 °C
                        7.E-07                                                                                           T=90 °C
                                                                                                                         T=250 °C
                        6.E-07                                                                                           T=500 °C
                        5.E-07

                        4.E-07

                        3.E-07

                        2.E-07

                        1.E-07
               ffective vapour diffusivity  [mE 0.E+00
                                                          0                0.2                        0.4        0.6     0.8         1

                                                                                                      Saturation  [-]


          Figure 5.24 The effective vapour diffusion coefficient for B-35 basalt 
          concrete at four different temperatures according to equations (5.59) and 
          (5.64) 



Another formulation of effective vapour diffusivity: 

                                                 1                                          j
                                                                                                
              gw                                                 va
                                                               D                       p
         v           = -                                  F               grad                   (5.60) 
          d                                                         j              g 
                                                                                                
                                                                                                
                                                    g
                                                              p                       p 

where  g =  (1-S) represents the ratio of material volume occupied by the 
gas phase (moist air), Dva/pg is the coefficient of free diffusion of vapour in 
air: 

                                                                                                         1.88
                                                                                                   
                     =                                         =                                       
              va
         D                                  v
                                   D a()                                0.217p                                   (5.61) 
                                                                                  atm                 
                                                                                                  o
                                                                                                       

and where the factor F, function of S accounts simultaneously for the 
tortuosity effects and the reduction of space offered to gas diffusion, is the 
one by Millington: 

         F = F (S )                                              4 / 3
                                                          = n              (1 - S )10/3  (5.62) 





 204


    Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




however, this formulation does not take into account the high temperatures 
effects, so it can be considered valid only in a range of temperatures close to 
the ambient ones. 
Now the important role played by the microstructure of the material in the 
diffusion of water vapour in the pores of concrete will be discussed. 

Influence of the pore structure on the vapour diffusion in concrete 
The pore space in concrete has a very complex inner structure, which 
influences a vapour diffusion process. The simplest way of taking it into 
account is introduction of the tortuosity factor,  , see e.g. [34],[35],[41], 
which mainly considers, as its name suggests, an increase of the averaged 
path of the diffusing molecules, which is caused by their `tortuous' course 
inside the pore space. The tortuosity factor is usually assumed to be 
constant in the whole range of moisture content, e.g. for concrete 0.4÷0.6, 
[15],[38]. Thus, the effective diffusion coefficient is defined as, [34],[41], 

       D (S ) =(1 - S )n  D ( p                          (5.63) 
         eff                           va       , g )

where Dva(,pg,) is diffusivity of vapour species in the air at the temperature 
 and pressure pg. 
But during diffusion of vapour in porous material with very narrow pores, 
e.g. in concrete, the number of collisions of water molecules with the solid 
matrix is non-negligible compared with the number of collisions with air 
molecules, thus Knudsen effects are of importance. This causes an 
additional decrease of the vapour diffusion coefficient, e.g. in a cylindrical 
pore of radius r by the factor of                              1           [15], where lm is the mean free 
                                                     1 + m
                                                          l 2r
path of water molecules in air, which depends upon temperature. If the 
pore-size distribution of the medium is known, then the effective vapour 
diffusion coefficient for different degrees of saturation may be evaluated 
theoretically from the formula proposed by Daian [15], 

                                               
                                                                    2
                                                          -
       D (S r )                          g           r         a
                                                               t                1
                     ( ) =  D     p                                                        d r  (5.64) 
         eff          k      va ( ,           )  ( r )                                       ( )
                                                                         1 + lm 2(r - t )
                                                                                      a
                                               rk

where  S(r ) is saturation corresponding to filling up all the pores with radii 
                k
smaller than rk, and d (
                              r) is volume fraction of the pores of radius between r 
and r+dr. 
Another formula, obtained by use of percolation theory, taking also into 
account the pore connections and their changes for different saturation 
degrees, was proposed by Chaube et al. [43], 





 205


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 



                                                                                          
                                                                                    
          D (S r )                                                   r V (r )
                                                     g               m         m                         1
                               ( ) =D          p                                   
                                                                                                                    d r
           eff                       k     va ( ,         )                                                              ( )
                                                                                     
                                                                                                                               (5.65) 
                                                                          c
                                                                          r                    1 + m
                                                                                                   l 2(r - t )
                                                                                                                  a
                                                                                          rk

where  V(r) is a fraction of the connected capillary pore space consisting of 
capillary pores with radii greater than  r,  rm is pore radius at which a 
maximum of the product of V(r) and r occurs, rc is critical pore radius, which 
can be determined from the inflection point of the saturation vs. r-1 curve 
[43]. 
Results obtained by use of equation (5.65) can be expected to be the most 
realistic ones (and, in general, considerable smaller then those obtained 
from the previous formulas), because of poorly interconnected air space in 
the pores of concrete, especially for higher degrees of saturation in materials 
with dominating narrow mesopores (i.e. approximately with the pore radii 
from 25 to 100 Angstroms), where capillary condensation occurs. 
The problem becomes more complicated, when temperature changes are 
taken into account. With temperature increase, the pore structure of 
concrete changes (see for example Figure 5.25, [42]), because of complicated 
dehydration and micro-cracking processes [42] , and concurrently the mean 
free path of water molecules increases [25],[42], what causes additional 
changes of the effective vapour diffusion coefficient. 
Thus, the influence of the Knudsen effect and changes of the pore 
connections (tortuosity factor) upon the vapour diffusion in concrete at 
various temperatures is taken into account in the present model by 
introducing into equation (30) the structure coefficient fK(,S), which can be 
evaluated from equation (5.64) or (5.65). 


                          0.05
                                                                                                              T=20 C
                          0.04                                                                                T=90 C
           on [-]                                                                                             T=250 C
                                                                                                              T=500 C
                     acti 0.03
                      fremu 0.02
                      volreo 0.01
                     P


                              0
                                     10         100                                 1000                 10000            100000

                                                                Pore radius [Angstrom]


          Figure 5.25 Change of the pore structure of the concrete at high 
          temperature. The pore size distribution curves of the B-35 basalt 
          concrete at four different temperatures 



 206


   Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




Examples of such calculations, using the pore-size distribution curves of the 
B-35 basalt concrete at four different temperatures, shown in Figure 5.25, 
done by use of the both equations (5.64) and (5.65), are presented in Figures 
5.26 a) and b). It can be seen, that in the saturation range, where vapour 
diffusion is of importance, Knudsen effects cause a significant decrease (2-10 
times) of the diffusive vapour flux. These effects will be even more evident 
for materials with greater volume fraction of very fine pores (i.e. close to the 
lower limit of mesopores), like for example HP and UHP concrete (fS 
0.001÷0.01), hence they should be taken into account, even for normal 
temperature range. Similar coefficient, called `attenuation factor', 
accounting for the restrictions imposed to vapour diffusion by the pore 
structure, was applied in [17] for wood and in [44] for concrete. In the both 
cases the value of this coefficient was assumed as equal to 10-5, but without 
sufficient justification and connection to the pore structure of the material. 
Considering the analysis presented in this section, the value of the 
coefficient assumed in [44] for concrete seems to be too small. 

                                                                  0.5

                                                                0.45                  T=20 C
                                                                                      T=90 C
                 [-]                                              0.4                 T=250 C
                                                                0.35                  T=500 C
                        e factor                                  0.3
                                    ctur
                                            tru 0.25
                                            S
                                                                  0.2

                                                                0.15
                                                                          0    0.1    0.2 0.3       0.4     0.5    0.6    0.7    0.8    0.9    1

                                                                                                    Saturation [-]

                                                                                                     a) 

                                                                 0.09

                                                                 0.08

                                                                 0.07
                                                  [-] 0.06
                                                                 0.05
                                                 e factor 0.04
                                                 ctur                                        T=20 C
                                                                 0.03
                                                         tru                                 T=90 C
                                                                S 0.02                       T=250 C
                                                                 0.01                        T=500 C
                                                                    0
                                                                          0    0.1    0.2    0.3    0.4     0.5    0.6    0.7    0.8    0.9    1

                                                                                                    Saturation  [-]

                                                                                                     b) 


        Figure 5.26 Structure coefficient of the B-35 basalt concrete at four 
        different temperatures: a) according to formula (5.64); b) according to 
        formula (5.65) 



 207


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




5.3.5 Dehydration 
The changes of the hydrated water content with temperature and the source 
term related to it: 

                                                                                                                      T
                                                                                                                        
               m
                                            =           m
                                                                              (T ) =                 m
                                                                                                       
                dehydr                                           dehydr                        (
                                                                                                                                  (5.66) 
                                                                                                            dehydr )
                                                  t                                         T
                                                                                                                       t

should be measured during laboratory tests. Some results of such tests, done 
for several types of concrete, are presented in [45]. For the temperature 
range below the critical point of water, the dehydrated water content 
  m
             (T ) increases, when temperature increases, approximately following 
      hydr
a step function where there is a sharp change between 200°C and 500 °C. 
For an ordinary concrete, in fact, the rise at the beginning of the 
phenomenon, i.e. when the hydration process stops around 105 °C, is rather 
small and only around 200 °C the dehydrated amount of water assumes 
significant values. 

               m
                                            = f m c f (T ) (5.67) 
                dehydr                            s


where  fs is stechiometric factor, m is ageing degree of concrete (between 0 
and 1), c is the cement content and f(T) is an function of temperature, i.e.: 

         if T < 105 C
                                             °                f (T ) = 0

                                                                               1 sin(3.1416/2(1 2exp( 0.004(T 105))))
                                                                                     +                            -               -       -
                                                                                                                                             
         if T                    105 C                        f (T )                                                                           
                    >                        °                            =                                                  2
          (5.68) 

The function f(T) is reported in the following figure: 

                                     1
                                0.9
                                0.8
                                0.7
                     [-]
                                0.6
                            f(T) 0.5
                                0.4

                            Function  0.3
                                0.2
                                0.1
                                     0
                                             0                     200                400         600             800             1000

                                                                                     Temperature [C°]


                                             Figure 5.27  dimensionless function of T of eq. (5.68) 



 208


      Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




5.3.6  Inner structure of concrete at high temperature: porosity and 
         intrinsic permeability 
The changes of porosity and intrinsic permeability with increase of the 
temperature and fluid pressure were measured for several types of concrete 
by Schneider and Herbst [42]. Their results showed, that the dependence of 
porosity on temperature can be approximated for concrete by a linear 
relationship: 

         n = o
                   n + n
                                      A (T -T ) (5.69) 
                                              0


where  An is a constant dependent on the type of concrete. For the 
experimental data for three types of the B35 - concrete [47], the following 
coefficients of equation (5.69) have been found by using the least square 
method: silicate concrete -  o
                                                   n =0.06,  An =0.000195 K-1; limestone concrete - 

 o
n =0.087,  An =0.000163 K-1; basalt concrete -  o
                                                                                 n =0.0802,  An =0.00017 K-1, 
see Figure 5.28. It should be reminded, that the porosity, n, and the 
                                                      
hydrated water source term,                                ( m
                                                                         , as well as the density of the solid 
                                                                hydr )
                                                     T
phase, s, must fulfil the solid mass balance equation (5.2). Thus, assuming 
that the first two parameters depend upon temperature according to 
equations (5.69) and (5.67), the change of solid density, caused by 
dehydration process, may be easily calculated through equation (5.2) in the 
temperature range (Tho,T), where Tho is temperature, at which dehydration 
process starts. Otherwise, knowing n and s as function of temperature from 
experimental tests, it is possible always using eq. (5.2), to evaluate the 
change of porosity related to dehydration process. 

                    0.2
                                      silicate-
                   0.18               concrete
                                      basalt-
                   0.16               concrete
                                      limestone-
                   0.14               concrete
             sity  [-]ro 0.12
             oP
                    0.1

                   0.08

                   0.06
                                 0     100          200           300          400     500     600

                                                      Temperature  [°C]



           Figure 5.28  comparison of the experimental values of total porosity for 
           three types of the B-35 concrete with approximate formula (5.69) 



 209


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




As far as the high performance and ultra high performance materials object 
of study in the framework of BRITE "Hiteco", mentioned at the beginning of 
this chapter, are concerned, several tests are carried out for measuring the 
porosity through mercurium porosimetry. 
Porosity characterization has been carried out in samples heated under 0% 
mechanical load as well as several samples heated under 30% load. Two 
different can be studied from  the porosity results: 
     ·  Total porosity of the material; 
     ·  Pore size distribution 
The evolution total porosity after heating the samples under 0% mechanical 
load are measured in different European laboratories on different samples 
of the same material and, because of this, the final experimental values do 
not coincide. 
In the Figure 5.29 they are summarised and also the linear interpolation 
according to equation (5.69) are reported. 
 

                              25



                              20                                                      C60
                                                                                      C60-SF
                                                                                      C70
                                                                                      C90
                              15
          [%]                                                                         RPC200
                                                                                      CRC
                 sity                                                                 Exper. values C60
                         roo 10                                                       Exper. values C60-SF
                         P                                                            Exper. values C70
                                                                                      Exper. values C90
                                                                                      Exper. values RPC200
                               5                                                      Exper. values CRC



                               0
                                    0    100    200    300     400    500    600

                                                 Temperature [°C]



                              Figure 5.29 experimental values measured for different types of HP 
                              and UHP concrete and linear approximation according to eq. (5.69) 



The HP concretes (C60, C60-SF, C70 and C90) show the same variation of 
porosity with temperature starting from different values, depending from 
the inner structure of the material which is more compacted for concrete 
with high strength value (the number indicated the final compression 
strength of concrete in MPa). 
UHP concretes are characterized by a thinner microstructure, hence they 
present lower values of porosity. In particular RPC200, a concrete in the 




 210


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




class of 200 Mpa, presents a constant porosity for the whole variation range 
of temperature. 
Regarding the total porosity and pore size distribution of specimens heated 
under 30% load, no relevant differences have been observed respect to the 
samples heated under 0% load (see Fig. 5.30 for C60 and C60-SF). There is 
only a small decrease of the number of pores between 0.01-0.1 mm ans 0.1-1 
mm in the case of 30% load test. In any case, the changes are not very 
significant. 
 


                           25



                           20



                           15
               [%]

                      sityroo 10
                      P
                                                                         C60-0%
                                                                         C60-SF-30%
                            5                                            C60-30%
                                                                         C60-SF-30%

                            0
                                    0    100         200          300           400     500    600

                                                            Temperature [°C]



          Figure 5.30 evolution of total porosity with temperature of C60 and 
          C60-SF specimens heated under 0% and 30% load 



Mercury porosimetry were carried out on various samples, in order to 
determine the influence of the heating time at a given temperature, after 
keeping the specimens during 1 or 5 days at a fix temperature (200°C for 
RPC200 or 400°C for C60 and CRC). Results are shown in table 5.3. no 
significant differences were observed with respect to those heated for 1 hour 
at the same temperature. 
 
            C60 
                                                                           RPC200 
                                                                                                      CRC 
400°C 1 hour                              13.59±1.54                                           8.41±1.24 
400°C 1 day                               13.90±0.27                                           9.90±1.61 
200°C 1 hour                                                              4.00±0.29                    
200°C 5 day                                                               4.62±1.20                    

             Table 5.3  influence of heating time on final values of total porosity 



 211


      Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




As an example of pore size distribution of such a kind of high performance 
concrete in Figure 5.31 that of C60 is shown. 
For further information see BRITE "Hiteco" final report. 

                                                              12
                                                                                20°C
                                                              10                105 °C
                                                                                200°C
                          on [-]                               8                400°C


                                                               6
                                    me fractiu
                                                               4

                                                  Pore vol
                                                               2


                                                               0
                                                                    10                    100         1000      10000    100000

                                                                                            Pore radius [Angstrom]



            Figure 5.31  pore size distribution for C60 high performance concrete at 
            various temperatures 



As well known, the intrinsic permeability, also called absolute permeability, 
is a material characteristic describing the penetration of gases or liquids 
through a porous material due to pressure head and it is generally 
determined by means of experimental tests performed using gas or water. 
Typical gases employed in these measurements are molecular oxygen and 
nitrogen. Theoretically, the intrinsic permeability coefficient of a given 
concrete should be the same regardless of whether a gas or a liquid is used 
in the tests. For gases, the compressibility as well as the viscosity of the gas 
must be considered in order to derive the coefficient of permeability: 

                    Ql                                                         2p
          kg = µ                                                                                       (5.70) 
                    tA ( p - p )( p + p )
                                                               1          2          1           2

where capillary flow has been considered as laminar and where: 
kg  = coefficient of permeability [m2] 
µ  = viscosity of the gas [Ns/m2] 
Q  = volume of gas flowing [m3] 
l      = thickness of penetrated section [m] 
A  = penetrated area [m2] 
p  = pressure at which Q is measured [Pa] 
p1  = pressure at entry of gas [Pa] 
p2  = pressure at exit of gas [Pa] 
t      = time [s] 


 212


      Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




Among liquids penetrating through concrete, water represents the most 
important fluid. In contrast to gases, liquids may be considered as 
incompressible, and eq. (5.70) may be then rewritten. If the viscosity is 
taken into consideration and a laminar flow is assumed, the coefficient of 
(water ) permeability is given by: 

                   Ql µ
            w
           k =                  (5.71) 
                   tA p
                        

where: 
kw  = coefficient of permeability [m2] 
µ  = viscosity [Ns/m2] 
Q  = volume of liquid flowing [m3] 
l     = thickness of penetrated section [m] 
A  = penetrated area [m2] 
p  = pressure difference [Pa] 
t     = time [s] 
As above mentioned, the coefficient of permeability measured using (5.70) 
and (5.71) represents a material characteristic. i.e. it depends only on the 
microstructure of the concrete, and is independent of the properties of the 
fluid. Instead, there are large differences between values obtained using 
water or gases. The gas values are consistently higher, the difference being 
greater at lower permeability coefficients, even if the material is practically 
dried. 
The difference may be explained by the theory of the slip in the flow of gas, 
whereby the gas close to a wall has a finite velocity. As a consequence, the 
quantity of gas flowing through a capillary is larger than would be predicted 
from Poiseille's formula, i.e. supposing a laminar flow. 
The gas slippage theory suggests that the gas flow will be affected by 
pressure which in turns influences the mean free path, and Klinkenberg 
derived an equation relating liquid and gas permeability coefficients (kl and 
kg) to the mean pressure (pm in atmospheres) at which the gas is flowing: 

                        kg
           k =                            (5.72) 
            l                b 
                  1 +             
                                 
                             m
                          p 

which is particularised by Bamforth for concrete through the following 
expression of the parameter b: 

                                   -8         0
                                         - .5227
           b = 1.635  10 k                          (5.73) 
                                         l

Figure 5.32 has been developed using equations (5.72) and (5.73) and 
illustrates the relationship between kl and kg for a range of mean pressure 
from 0.1 to 5 MPa. 




 213


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




For typical structural concrete, which is likely to have an intrinsic 
permeability in the range 10-19 to 10-17 m2, gas permeability values may be 
one to two orders of magnitude higher. These differences are more evident 
for concrete with a narrow inner structure, like high and ultra-high 
performance concrete. Hence, for such a kind of concrete it is of importance 
to consider the effect of slippage measuring intrinsic permeability because 
for these material the final values can be three orders of magnitude 
different. 
 

                                1.E-14


                                1.E-15


                                1.E-16


                                1.E-17
            ]2                                    0.1 Mpa                                                0.1 MPa
                                1.E-18
                   [m                                                                                    0.5 MPa
                         k g                                                                             1 MPa
                                1.E-19                                                                   2 MPa
                                                   5 Mpa                                                 3 MPa
                                1.E-20                                                                   4 MPa
                                                                                                         5 MPa
                                1.E-21            line of equality                                       Equivalent


                                1.E-22
                                    1.E-22    1.E-21     1.E-20       1.E-19    1.E-18     1.E-17    1.E-16    1.E-15    1.E-14

                                                                                kl [m2]



           Figure 5.32 theoretical relationships between water permeability and 
           gas permeability derived using Klinkenberg's equation (5.72) for different 
           mean gas pressure 



The problem concerning the measure of absolute permeability is decisively 
more complex if the temperature is not constant, in particular at high 
temperature, i.e. above 100°C. 
In such a case, many factors can influence the results. Usually, the measure 
is the result of a "residual" water or gas permeability test conducted at 
ambient temperature after a heat treatment. 
Heating concrete from ambient temperature to 120°C increases k by up to 2 
orders of magnitude. Moreover, drying (e.g. in the range 120-600°C) causes 
internal structural damage and dehydration of the cement, resulting in (a) 
an increase in k value, and (b) k values which are time dependent initially 
on re-wetting. In fact, the dehydration caused by drying at high 
temperature is partially reversed on exposure to moisture, resulting in a 



 214


   Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




reduction of the k values by 1 or 2 orders of magnitude. However, the 
resulting values of k remain high when compared to a material which has 
not been heated, e.g. up to 6 orders of magnitude. The damage increases 
with increase of temperature, thus resulting in progressively increasing of k 
with temperature increase. The curing treatments are of importance, too. In 
fact, with particular curing during heating of concrete it is possible to 
minimise damage degree, resulting in a lower k increase. Hence, damage 
plays a role really important and should be taken into account. For the 
evaluation of intrinsic permeability temperature dependence, different 
approaches can be followed. A phenomenological approach is usually applied 
for description of changes of concrete physical properties during complex 
hygro- thermal and mechanical phenomena at high temperature, e.g. [20]. It 
means that all these changes are expressed as a function of temperature, 
moisture content and gas pressure, i.e. physical quantities measured 
directly during experimental tests. However results of such tests are 
strongly dependent upon the form and dimensions of a test sample, as well 
as physical conditions during experiment. Hence application of the results of 
these tests for prediction of concrete behaviour in conditions that differ 
significantly from the experimental ones is rather questionable. 
Because of this there is a need for mechanistic (structural) mathematical 
models which assume certain physical models of the phenomena analysed. 
As above mentioned, significant increase of concrete intrinsic permeability 
at high temperature is mainly generated by arising micro-cracks and by 
changes of material inner structure, as well as by crack-opening due to high 
gas pressure values. As a result, it depends not only upon temperature, 
moisture content and gas pressure, as assumed for a phenomenological 
approach [20], but also upon a degree of cracks development, which may be 
described by use of damage parameter D. 
In Reference [42] Schneider and Herbst presented results of their research 
on changes of inner structure and permeability, resulting from high 
temperature action, for four types of concrete. Results of these tests, 
concerning effect of temperature and gas pressure on concrete intrinsic 
permeability, have been approximated in References [32],[46] by use of a 
formula of phenomenological type see Fig. 5.33: 

                                                AP
                                      p
                       A ( 
                             - )                
      k =                                    
             o
             k 10                          g
                       T      o                 (5.74) 
                                               
                                           o
                                           p 

where  k0 is intrinsic material permeability at reference temperature 0= 
293.15 K and reference pressure p0= 101325 Pa, AT and AP are material 
dependent constants. 
Bary [47] measured changes of concrete intrinsic permeability caused by 
mechanical material damaging at ambient temperature. He proposed the 
following formula describing this phenomenon: 

      k = o
             k 10A D
                        D     (5.75) 



 215


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




where material constant AD was evaluated as equal to 4. 
 

                                                     1.6E-14

                   ] 1.4E-14                                                 1 atm
            2 m                                                              2 atm
                    [ 1.2E-14
                    k                                                        3 atm
                         y                                                   5 atm
                              ilit 1.0E-14
                              ba 8.0E-15
                              rme
                                      pe 6.0E-15
                                     ic
                                                     4.0E-15
                                           rins
                                                   Int 2.0E-15

                                                    0.0E+00
                                                               293.15 343.15 393.15 443.15 493.15 543.15 593.15 643.15

                                                                                                 Temperature  [K]



           Figure 5.33  dependence of the intrinsic permeability of the B-35 silicate 
           concrete upon temperature and pressure according to approximate 
           formula (5.74) 



During heating of concrete at high temperature complex physical and 
chemical processes take place, resulting in changes of inner structure, 
micro-cracks development and porosity increase, [30],[31]. They cause an 
increase of intrinsic permeability [42]. There is still lack of sufficient 
experimental research to form a basis of a mathematical model of these 
complex phenomena. However one can expect that a joint effect of 
temperature, gas pressure and material damaging (crack development) on 
the intrinsic material permeability, K, may be described by a combination of 
the previous two formulas, (Fig. 5.34),  

                                                                                            p
                                                                                       A
                                                                             p
                                                               A (T T
                                                                    - )                
         k =                                                                        
                                                     o
                                                    k 10                          g               A D
                                                               T     o                10 D
                                                                                      
                                                                                  o
                                                                                  p                                      (5.76) 

where  AT,  AP and AD are material constants. It should be noted that 
coefficients AT and AP in Eq. (5.76) should be lower than coefficients in (5.74) 
because phenomenological model (5.74) takes indirectly into account also 
influence of concrete cracking which may be described by use of damage 
parameter D, [48]-[50]. Parameter AP in (5.76) has a clear physical 
interpretation, because it describes effect on the material permeability 
increase of crack opening caused by pressure rise. A value of parameter AD 
in (5.75) and (5.76) is dependent on the type and dimensions of the cracks 
developed in concrete matrix. Hence for the hygro-thermal and mechanical 



 216


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




phenomena accompanying concrete heating one can expect lower value of AD 
than for the classical concrete strength tests Eq. (5.75). 
 


                         1E-13



                         1E-14
          ]2

                Ko [m    1E-15

                                                                                                                                                  600
                          1E-16 5                                                                                                            450
                                           5                                                                                                        T [K]
                                     0.                5
                                                1.                5
                                                            2.               5
                                                                       3.               5
                                                                                  4.               5
                                                                                             5.               5                            300
                                                                                                        6.                                                   -14--13
                                                                                                                   7.    58.    5
                                                      Pg/Patm  [-]                                                                   9.                      -15--14
                                                                                                                                                             -16--15     

          Figure 5.34 
                                           intrinsic permeability as function of Pg, T and 
          damage,equation (5.76) with Ak=0.0025, Bk=0.368, AD=2. 



In order to confirm the importance of the role played by the 
thermomechanical damage on intrinsic permeability and, from a more 
general point of view, on the inner microstructure of concrete, in the 
following a description of the results of an examination of different concretes 
objects of study in the framework of BRITE "Hiteco" research project on 
fluorescence microscopy, will be reported. 
This type of microscopy provide the necessary information concerning 
aspects such as density of the paste, homogeneity and the development of 
cracks. The description is based on the changes that appear in each type of 
concrete due to the exposure to different temperatures. 
The concrete chosen are C60, C60-SF and CRC representing typical high 
performance concrete, and an its variation with silica fume, and a typical 
ultra high performance concrete. 
In this description the crack width is used as class references: 
     ·  wide cracks: crack width over 0.1 mm, 
     ·  fine cracks:  crack width 0.01-0.1 mm, 
     ·  microcracks: crack width below 0.01 mm. 
 Orientation of the cracks - as could be expected for microcracks developed 
due to external loading- is not observed in any thin section. The cracks are 
divided into three categories: 



 217


  Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




   ·  debonding cracks (cracks in the interfacial zone between aggregate 
      and paste), 
   ·  paste cracks (cracks through the paste matrix), 
   ·  aggregate cracks (cracks through aggregates) 
All photos have the same scale: 1 cm equals 270 µm 
                                                 





                                                                                           
                                              a) 
                                                 





                                                                                           
                                              b) 


        Figure 5.35  a) C60 reference. Part of thin section that shows high 
        capillary porosity in the transition zone between aggregates and paste. 
        Dark areas show low capillary porosity; b) C60 at 500 °C. Pattern of 
        paste microcracks and the fine and wide debonding cracks. 




 218


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




C60, C60-SF: 
The photo in Fig. 5.35 a) shows that the capillary porosity is high in the 
transition zone between aggregate and paste, particularly around the 
largest aggregates. The sample is in that respect not completely 
homogeneous. There is a low content of microcracks. They occur as paste 
and as debonding microcracks.. 
The sample dried at 105°C has a higher capillary porosity. The paste is 
homogeneous and there is no sign of cracks. 
Exposed to 500°C cracks occur as shown in Fig. 5.35 b). From the 
observations of the thin sections no change in the capillary porosity is 
apparent, but there is a slight improvement in homogeneity. It is unlikely 
though, that this improvement in homogeneity is due to the change in 
temperature. Rather, the difference should be attributed to the variation 
which can be expected when performing measurements on different 
specimens from the same batch. The development of microcracks most likely 
initiates before 500°C has been reached, but the cracks are only evident in 
the sample exposed to 500°C.The cracks occur in two forms: as paste 
microcracks and as fine and wide debonding cracks. A few examples of 
aggregate cracks occur but these are not common. C60 was also exposed to 
700°C, but the test sample was severely cracked and it was not possible to 
prepare a thin section. 
 In the C60-SF sample it is evident that an air entraining agent has been 
used. The reference sample shows circular air voids with a range of 
diameters. The paste shows inhomogeneities which are evident from the 
variations in capillary porosity (Fig. 5.36 a)). Examples of debonding 
microcracks occur, but are very rare. 
The sample dried at 105°C has a higher capillary porosity than the 
reference and the paste is homogeneous and show no evidence of 
microcracks. 
When heated to 500°C air voids start to agglomerate. Fine and wide 
debonding cracks and paste microcracks are extended by the assimilation of 
some of the air voids and a few aggregate cracks occur. There is no change 
in the capillary porosity and the paste is homogeneous. This marked 
agglomeration leads to large irregular areas of voids when the concrete is 
exposed to 700°C, i.e. it is very porous and a marked reduction of paste 
density is evident. At this temperature aggregate cracking is common and 
wide debonding cracks occur, Figure 5.36 b). 
 
CRC: 
The reference sample has some few, very thin paste microcracks. The paste 
is generally very dense and homogeneous and microcracking is rare in the 
reference sample and also in the sample dried to 105°C. The sample exposed 
to 105°C is similar to the reference sample concerning homogeneity and 
density. The length of microcracks has increased (Fig. 5.37a)) up to 2 mm 
compared to 0.5 mm in the reference sample. 



 219


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




Microcracks could be expected for a concrete as dense as this. Early 
shrinkage (i.e. due to drying) is normally greater the larger the cement 
content of the mix. Without pre-drying, the concrete might avoid a 
development of microcracks. 
 





                                                                                             
                                                 a) 
                                                    





                                                                                                  
                                                 b) 


           Figure 5.36  a) C60-SF reference. Pattern of air voids and a slightly 
           inhomogeneous paste. No cracks; b) C60-SF at 700 °C. Agglomerated air 
           voids causing large irregular voids. Mark the higher capillary porosity 
           than in (a) and the lack of higher capillary porosity in the paste-
           aggregate interface. Cracks in aggregate are shown at the lower left of 
           the photo. 





 220


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




After exposure to 700°C a decrease in the capillary porosity occur as clearly 
visble in Figure 5.37 b) which also shows severe cracking including all types 
of cracking and cracking widths 
 





                                                                                               
                                                 a) 
                                                    





                                                                                               
                                                 b) 


          Figure 5.37  a) CRC at 105 °C. This sample has a low capillary porosity. 
          Paste cracks a bit longer than for the reference sample are visible. Few 
          circular air voids are present; b) CRC at 700 °C. Photo shows severe 
          cracking system with all kind of cracks including cracks in quartz 
          aggregates. The capillary porosity is very low. 





 221


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




Observations of the reference samples from each type of concrete reveals 
some differences in homogeneity and there are, of course, differences in the 
capillary porosity due to different mix compositions. 
Especially C60 shows degrees of paste inhomogeneity. C60-SF shows a 
lower degree of inhomogeneity and finally CRC generally has homogeneous 
pastes. The reason for the inhomogeneities can be inadequate mixing. 
In the reference samples of C60, C60-SF and CRC microcracks occur, but 
are not a common feature. The microcracks in  the reference samples may 
have developed during thin section preparation. 
The reference samples of C60 and C60-SF show high capillary porosity in 
the paste-aggregate interface. The content of, silica fume in C60-SF has not 
been sufficient for improving the microstructure by densifying the porous 
paste-aggregate interfacial zone as could be expected. 
For both C60 and C60-SF the porosity increases when exposed to 
temperatures above 105 °C. In this range the loss of chemically bound water 
causes changes in the microstructure of the cement paste, e.g. shrinkage. 
Also the chemical decomposition cause significant changes: The irregular 
voids and the higher porosity observed above 500 °C can possibly be due to 
the dehydroxylation of Ca(OH)2, and the general loss of crystallinity, 
although rehydration products may occur during subsequent cooling, 
refilling the voids if water is present. 
C60 was severely cracked after heat treatment at 700 °C. The sample 
containing silica fume, C60-SF did not suffer to the same degree. After 
treatment at 700 °C it shows a very porous paste with large voids and 
aggregate cracking of quartz. Heating at 700 °C causes an endothermal 
reaction including a complete dehydroxylation of Ca(OH)2 (in a range from 
440 to 580 °C) and a partial decarbonatisation of CaCO3. 
The cause of the severe cracking of C60 at 700 °C is not easily discerned. 
The lack of silica fume compared to C60-SF should be an advantage; the use 
of silica fume leads to instability above 300 °C. Hertz (1992) concludes that 
concrete densified by means of silica fume is more likely to explode. On the 
other hand, however, the use of silica fume and fly ash combined with the 
lime filler in the cement could minimise the content of Ca(OH)2 in the 
C60-SF concrete. Release of water from Ca(OH)2 at high temperatures can 
cause severe cracking at later ages, when CaO is rewetted and increases in 
volume. 
The influence of the air entraining agent in C60-SF is unlikely to be the 
cause, as the air entrainment should have more influence with regard to 
prevention of spalling than with regard to the formation of residual 
cracking. 
Another aspect is the various types of aggregates used. Calcareous 
aggregates used in C60 should be stable until 800 °C. At this stage the 
limestone aggregates calcinate causing expansion and cracking (Metha, 
P.K., 1986). Possibly the calcination started at 700°C, as the difference in 
aggregate type seems to be the major difference between the two C60 
concretes. 



 222


   Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




In C60-SF, quartz-aggregates undergo a phase transformation leading to a 
volume increase of 1 %. The aggregate cracking at 700 °C (and 500 °C) is a 
result of aggregate expansion and the decomposition of paste. 
Since CRC showed spalling when heated to 500°C, no thin sections were 
prepared. A CRC sample dried for one day at 105°C before heated to 700°C, 
however, did not present a spalling problem. 
This implies that one of the main reasons for the lack of fire resistance and 
the degree of spalling at high temperatures could be the water content. Not 
the chemically bound water, because the chemically bound water is only 
partially lost at 105°C , but the adsorbed and interlayered water. When 
adsorbed and interlayered water is lost (65-100°C) from predrying, CRC is 
resistant to spalling at 700°C. 
The reference sample and the CRC sample dried at 105°C has paste cracks. 
For more detailed information see the BRITE "Hiteco" final report [51]. 

5.3.7  Importance of different physical mechanisms of moisture 
      transfer at high temperature 
Using the material data presented in this chapter, it is possible to estimate 
the relative importance of specific physical mechanisms of moisture transfer 
in concrete at different temperatures. In order to do this, it is necessary to 
relate all the moisture fluxes to the same gradients of state variables, e.g. 
capillary pressure, gas pressure and temperature. In this way, by comparing 
the coefficients corresponding to the gradients of the same state variable, it 
is possible to evaluate their relative magnitude, assuming that other state 
variables remain constant. Below, an example of such an analysis is 
presented for a concrete with intrinsic permeability K=3.210-17 m2, by using 
the sorption isotherms calculated from the approximated formula proposed 
by Bazant and Thongutai [30],[31] and other material parameters, as 
presented in the previous sections. 
The following transport phenomena are related to the capillary pressure 
gradient: vapour diffusion, bound water diffusion and flow of capillary 
water. Figure 5.38 shows the dependence of these fluxes upon the relative 
air humidity, for the case of unit capillary pressure gradient at three 
different temperature values, while the gas pressure remains constant and 
equal to the atmospheric pressure value.  
The gas pressure gradient causes the Darcian flows of vapour and capillary 
water, as well as vapour diffusion. The dependence of these fluxes upon the 
air relative humidity, for the unit gas pressure gradient, for the three 
different temperatures (the same as for the previous case) and constant 
capillary pressure values, corresponding to the given temperature and 
relative humidity values, is presented in Figure 5.39. 
When capillary pressure and gas pressure remain constant, the temperature 
gradient itself causes only vapour diffusion (sometimes called also thermo-
diffusion). However, when large temperature gradients occur in material, 
like for example during a rapid heating of a concrete wall, the vapour 
thermo-diffusion transport is of importance. 



 223


Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




                                                                                                           1E-10

                          c                                                                                1E-11         T=25°C
                                                                                                           1E-12

                                                                                                           1E-13                    Capillary water flux

                                                                                                           1E-14

                                                                                                           1E-15
                                                                   s related to grad p                                                                  Water vapour diffusion
                                                                                                           1E-16

                                                                                                           1E-17
                                                                                          ass flow         1E-18
                                                                                                      M                                            Adsorbed water diffusion                 A
                                                                                                           1E-19
                                                                                                                    0       0.2             0.4             0.6        0.8             1

                                                                                                                                      Relative humidity  [-]


                                                                                                           1E-10

                     c                                                                                     1E-11         T=100°C

                                                                                                           1E-12

                                                                                                           1E-13                   Capillary water flux

                                                                                                           1E-14

                                                                                                           1E-15
                                s related to grad p                                                                                                          Water vapour diffusion
                                                                                                           1E-16

                                                                                                           1E-17
                                                       ass flowM 1E-18                                                                     Adsorbed water diffusion                         B
                                                                                                           1E-19
                                                                                                                    0       0.2             0.4             0.6         0.8            1
                                                                                                                                      Relative humidity  [-]


                                                                                                           1E-10

                c                                                                                          1E-11
                                                        p                                                                T=360°C
                                                             rad 1E-12                                                                      Water vapour diffusion
                                                              go 1E-13
                                                              tdte 1E-14
                                                             lae 1E-15
                                                             s rw 1E-16                                                            Capillary water flux             Adsorbed water
                                                             lo                                                                                                         diffusion
                                                                                                           1E-17

                                                                   Mass f 1E-18                                                                                                             C
                                                                                                           1E-19
                                                                                                                    0       0.2             0.4             0.6         0.8            1

                                                                                                                                      Relative humidity  [-]



      Figure 5.38 Comparison of the mass fluxes, related to the capillary 
      pressure gradient, corresponding to different moisture transfer 
      mechanisms, at three different temperature values (grad pc=1 Pam-1, 
      pg=patm). A) T=25°C, B) T=100°C, C) T=360°C 




 224


Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 



                                                                                            1.E-08

                 g                                                                          1.E-09         T=25°C
                                          ad p 1.E-10
                                           gr                                                                                                     Capillary water flux
                                                 o
                                                      d t 1.E-11
                                                      tela 1.E-12
                                                      e                                                           Water vapour diffusion
                                                                                            1.E-13
                                                           lows r 1.E-14
                                                                                            1.E-15                                              Water vapour advection
                                                            Mass f                                                                                        flux                      A
                                                                                            1.E-16
                                                                                                      0        0.2             0.4              0.6           0.8              1

                                                                                                                          Relative humidity  [-]



                                                                                            1.E-08

            g                                                                               1.E-09         T=100°C
                           ad p 1.E-10
                            gro 1.E-11                                                                        Water vapour diffusion
                           d ttela 1.E-12
                           e
                                                                                            1.E-13
                                lows r                                                                     Water vapour advection flux
                                                                                            1.E-14                                               Capillary water flux

                                          Mass f 1.E-15                                                                                                                             B
                                                                                            1.E-16
                                                                                                      0        0.2             0.4              0.6          0.8               1

                                                                                                                          Relative humidity  [-]



                                                                                            1.E-08
                                                                                                                      Water vapour diffusion
                      g                                                                     1.E-09
                                                                      ad p 1.E-10
                                                                       gr                                                   Water vapour advection flux
                                                                             o 1.E-11
                                                                             d ttela 1.E-12
                                                                             e

                                                                                            1.E-13
                                                                                  lows r 1.E-14                                                             Capillary water
                                                                                                                                                                      flux
                                                                                            1.E-15
                                                                                  Mass f                   T=360°C                                                                  C
                                                                                            1.E-16
                                                                                                      0        0.2             0.4              0.6          0.8               1

                                                                                                                          Relative humidity  [-]



     Figure 5.39  Comparison of the mass fluxes, related to the gas pressure 
     gradient, corresponding to different moisture transfer mechanisms, at 
     three different temperature values (grad pg=1 Pam-1,  pc=const). A) 
     T=25°C, B) T=100°C, C) T=360°C 



 225


   Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




From the presented figures and considerations, based on the model 
proposed, it may be deduced that, generally speaking, for concrete, with any 
temperature increase a relative importance of the vapour diffusion and 
Darcian flow increases (particularly, when high gas pressure gradient 
occurs), while the liquid water flow is of importance only for the relative 
humidity values close to one and temperature range below approximately 
140°C. 
The bound water diffusion is of importance only for a temperature range 
below approximately 50°C and for very low relative humidity values of the 
air. Such general conclusions could be expected even from a qualitative 
analysis of the physical phenomena involved in moisture transfer at high 
temperature, but here also some quantitative estimations, ensuing from the 
presented model and material data, have been obtained. They are also, in 
general, in accordance with widely used one-phase phenomenological models 
of concrete at high temperature, e.g. [30],[31], except of a smaller, as 
compared with fenomenological formulation [30],[31], range of temperature 
and relative humidity, where the bound water flow is of importance. 

5.3.8  Thermal properties of concrete at high temperature: thermal 
       conductivity and thermal capacity 

Thermal conductivity 
The effective thermal conductivity of partially saturated concrete depends 
upon the degree of saturation with water and temperature, [21], 

                                 4      w
                                       n  S 
                 =  (T )1+                        (5.77) 
           eff       d     (1 - n ) s
                                               
                                                    

where  d (T) is thermal conductivity of a dry material, which temperature 
dependence may be usually approximated by a linear relationship [45], 

        =  [1 + A (
                                T - T )  (5.78) 
           d        do                        o ]

An example of the effective thermal conductivity for concrete with the 
following data: do =1.67 W/(mK), A=0.0005 °C-1 and To=25 °C is presented 
in Figure 5.40. As far as high performance and ultra high performance 
concrete are concerned, during "Hiteco" project numerous tests for 
measuring thermal properties of these materials have been carried out. 
Unfortunately, all tests have been performed in non thermodynamic 
equilibrium conditions, in the sense that the specimens were heated 
following certain heating rate and then kept for a short period, in the order 
of few hours, at reached temperature. The duration of the plateau was not 
enough for the achieving of steady state, i.e. concrete did not lose the whole 
quantity of water corresponding to plateau temperature because it has been 
cooled down when it was still loosing water. For instance, the plateau at 
200°C is rather short (1 hour) and not all removable water had gone away. 



 226


     Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




This is the main reason for the strange tendency shown by the values of 
thermal conductivity measured experimentally (Fig. 5.41). In fact, it 
possible to observe a non monotone decreasing with temperature as 
expected (because a deterioration of the inner structure with heating). 
 

                                                                               1.7
                                                                                                                                        Pc= 0 [Pa]
                                                                                                                                        1.00E+08
                                                                               1.6                                                      2.00E+08
                                                                                                                                        4.00E+08
                                                                               1.5                                                      6.00E+08
                                                                                                                                        5.00E+07

                                                                               1.4


                                                                               1.3


                                                                               1.2
              Thermal conductivity  [W/(mK)]

                                                                               1.1
                                                                                      20    70     120       170       220      270     320         370

                                                                                                          Temperature  [oC]


          Figure 5.40  dependence of thermal conductivity of an ordinary concrete 
          upon temperature and capillary pressure according to equation (5.77) 
          (assuming S(pc,T) from eq. (5.49), do=1.67 W/(mK), A =0.0005 K-1 
          so=2590 kg/m3 and n=0.1 

 

                                                                               3.5

                                                                                 3

                                                                               2.5                                                                         C60
                                                                                 2                                                                         C60-SF
                                                                                                                                                           C70
                                                                               1.5                                                                         C90
                                                                                                                                                           RPC200
                                                                                 1                                                                         RPC/AF
                                                                                                                                                           CRC
                                                                               0.5
                                                Thermal conductivity [W/mK]

                                                                                 0
                                                                                      0     100     200         300           400      500      600

                                                                                                          Temperature [°C]



          Figure 5.41 high and ultra high performance concrete experimental 
          results of thermal conductivity measurements 



 227


     Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




In the model proposed in the previous chapters one of the fundamental 
hypotheses is the thermodynamic equilibrium, thus, from the point of view 
of our model, these experimental results would not be directly applicable, 
because we do not know exactly the quantity of exiting water still present in 
the specimen at the moment of the measurement. 
 
                               THERMAL CONDUCTIVITY [W/mK] 
                                                                                                                            
                                                                             Temperature 
          Material                      20                     105          105                    200             400          600 
          C60                           2.21 2.04 1.79 1.95 1.54 1.17 
          C60-SF                        1.98 1.75 1.67 1.75  1.4  1.02 
          C70                           1.65 1.45 1.33 1.42  1.4  1.03 
          C90                           1.67  1.4  1.31 1.26 1.17 0.95 
          RPC200                        2.32 2.35 2.39 2.35                                                                       
                                                                                                                                   
          RPC/AF                        2.36 2.44 2.26 2.43 1.98 1.76 
          CRC                           3.05 2.98 2.72 2.54 2.03 2.16 


               Table 5.4  experimental data of thermal conductivity for high and ultra 
               high performance concrete [51] * stands for dry 



Taking into account the limitations due to the conditions during 
experimental tests, it is however possible to deduce a reasonable thermal 
conductivity law for these materials from the date shown in Table 5.4. 
In fact, the data at 105°C in dry condition, i.e. steady state, it is utilizable 
for determining the thermal conductivity of "dry" concrete, while, for this 
aim, a second point can be obtained considering the value at 600°C. This 
value has not been determined at steady state, but one can suppose at that 
temperature the quantity of water (liquid or vapour) inside the pores of 
material is really small because at 600°C the kinetics are much higher and, 
moreover, the corresponding plateau is longer (4 hours). Thus we can 
consider concrete in quasi-dry conditions at 600°C. 
In such a manner it is possible to calculate a linear relationship for d 
similar to that equation (5.78) and then to consider the influence of the 
water present into the voids of concrete trough a direct thermal 
conductivity-temperature and capillary pressure relationship, like the 
following, which can be used instead of (5.77),: 

                                                    c                               c              c
                                          p                                  p - p 
                     =  ( c
                               p ,T ) =                 (
                                                        aT + b )                                o
                                                                            -                          (
                                                                                                                          (5.79) 
                                                                                                      cT + d )
               eff      eff                         c                                         c         
                                               o
                                               p                                        o
                                                                                         p              

where  co
          p  is a limit value for capillary pressure corresponding to a condition 
in which the saturation is practically zero. 





 228


                                                                                                                Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




For a value of 6e08 Pa the desorption curves, for an ordinary as well as a 
high performance concrete, give a value close to zero (black bold line in Fig. 
5.42 and coefficients a and b in eq.(5.79)).  
Coefficients c and d are modelling the variation of thermal conductivity with 
capillary pressure and determine the line corresponding to full saturation 
S=1 (pc=0). 
As clearly visible from Figure 5.42, uhp concretes (CRC, for example), which 
are fibres reinforced, present a thermal conductivity much higher than an 
ordinary or a high performance concrete because of their higher 
compactness and lower porosity and because of the presence of steel fibres. 
 

                                                                                                                2.4                                                                                                                                                                     2.2
                                                                                                                                                                 0.00E+00                                                                                                                                                            0.00E+00
                                                                    k] 2.2                                                                                       1.50E+08         k]                                                                                                                                                 1.50E+08
                                                                                                                                                                                                                                                                                         2
                                                                    /m                                                                                           3.00E+08               /m                                                                                                                                           3.00E+08
                                                                          W                                                                                                                   W
                                                                                [                                    2                                           4.50E+08                           [                                                                                                                                4.50E+08
                                                                                     ty                                                                          6.00E+08                                                                                                               1.8
                                                                                                                                                                                                         ty                                                                                                                          6.00E+08
                                                                                           ivi 1.8                                                               Exper. values                                 ivi                                                                                                                   Exper. values
                                                                                                                                                                                                                                                                                        1.6
                                                                                                                1.6
                                                                                            conduct                                                                                                                    conduct 1.4
                                                                                                       al 1.4                                                                                                         al

                                                                                                                1.2                                                                                                                                                                     1.2          C60-SF
                                                                                                       Therm                      C60                                                                                       Therm

                                                                                                                     1                                                                                                                                                                   1
                                                                                                                          20       120    220     320     420          520                                                                                                                     20     120      220    320     420           520
                                                                                                                                           Temperature [°C]                                                                                                                                                    Temperature [°C]



                                                                    1.7                                                                                                                                                                                                                 3.2
                                                                                                                                                               0.00E+00                                                                                                                                                              0.00E+00
                                                                    1.6
     k]                                                                                                                                                        1.50E+08                                                              k]                                                                                              1.50E+08
                                                                                                                                                                                                                                                                                          3
           /m                                                                                                                                                  3.00E+08                                                                    /m                                                                                        3.00E+08
                                                                    1.5
                 W                                                                                                                                                                                                                               W                                                                                   4.50E+08
                       [                                                                                                                                       4.50E+08                                                                                [
                            ty                                                                                                                                                                                                                                                          2.8                                          6.00E+08
                                                                    1.4                                                                                        6.00E+08                                                                                     ty
                                  ivi                                                                                                                          Exper. values                                                                                      ivi                                                                Exper. values
                                                                    1.3                                                                                                                                                                                                                 2.6

                                                                    1.2
                                          conduct                                                                                                                                                                                                                         conduct 2.4
                                                     al                                                                                                                                                                                                                  al
                                                                    1.1
                                                                                                                                                                                                                                                                                        2.2           CRC
                                                           Therm                                                1                C90                                                                                                                                           Therm

                                                                    0.9                                                                                                                                                                                                                   2
                                                                                                                     20           120     220    320     420           520                                                                                                                     20     120      220     320         420       520
                                                                                                                                          Temperature [°C]                                                                                                                                                      Temperature [°C]



                                                                                                                                Figure 5.42  thermal conductivity as direct function of temperature and 
                                                                                                                                capillary pressure for high and ultra high performance concrete provided 
                                                                                                                                in "Hiteco" project 



Specific heat and thermal capacity 
The thermal capacity of partially saturated concrete can be expressed as a 
combination of the thermal capacities of its constituents, [11],[18],[52], 





 229


          Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




     C
              = (1 - n ) s                                                    w
                                                     C + n S  C                           + (1 - S )( g                   gw                       
          p                                                ps                       pw                     Cpga +  (Cpgw -Cpga ))
                                                                                                                                                    
 (5.80) 

where thermal capacities Cp depend upon the temperature of the material. 
A temperature dependence of the thermal capacity  (C)ps for the solid 
skeleton may be approximated by a linear relationship, [45], i.e. 

                      C
                                                  = C
                                                                [
                                             ps           pso 1 + A (T - T ) ]
                                                                           c               o       (5.81) 

where  Cpso  is the thermal capacity of the solid skeleton at the reference 
temperature To, Ac - a coefficient. 
An example of the thermal capacity for a partially saturated concrete (Cpso 
=940 J/(K.kg), so=2590 kg/m3,  n=0.1), obtained by using saturation - 
capillary pressure relationship from [30],[31], is shown in Figure 5.42. 
 
                                             2.45E+06
                                                                                                                                  Pc= 0 [Pa]
                                             2.40E+06                                                                             5.00E+07
                                                                                                                                  1.00E+08
                      K] 2.35E+06
                 3                                                                                                                2.00E+08
                      m 2.30E+06                                                                                                  4.00E+08
                      ty [J/                                                                                                      6.00E+08
                                             2.25E+06

                                             2.20E+06
                                 capacial 2.15E+06
                                m
                                             2.10E+06
                                     Ther
                                             2.05E+06

                                             2.00E+06
                                                          20          70             120           170    220        270          320         370

                                                                                                Temperature  [oC]


                      Figure 5.43 dependence of thermal capacity of an ordinary concrete 
                      upon temperature and capillary pressure according to equation (5.77) 
                      (assuming S(pc,T) from eq. (5.49), Cpso=940 J/(kgK), so=2590 kg/m3 and 
                      n=0.1 



It is possible to give a direct dependence of thermal capacity, or specific 
heat, of partially saturated porous media, from the state variables T and pc, 
without considering a combination of the thermal capacities of single 
constituents, i.e.: 

                      C
                                                  = C
                                                          (
                                             p            p T, c
                                                                      p ) (5.82) 





 230


        Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 




For example, in the case of high and ultra-high performance concrete whose 
experimental values present the same problems than thermal conductivity, 
a relationship similar to (5.79) can be used for the (apparent) specific heat of 
concrete: 

                                      c                             c              c
                                 p                            p - p 
             =  ( c
                      p ,T ) =            (
                                          aT + b )                              o
                                                         -                             (
                                                                                                         (5.83) 
                                                                                      cT + d )
                                      c                                       c         
                                      o
                                      p                                  o
                                                                         p              

The graphs in Figure 5.44 show the specific heat for some concretes while 
Table 5.5 summarises the experimental data for these materials. 
 
                                      SPECIFIC HEAT [J/KgK] 
                                                                                                                        
                                                                         Temperature 
             Material                      20          105                105*                    200          400          600 
             C60                           1095 1005 892  923 741 
                                                                                                                            796 
             C60-SF                        1063 1006 832  843 806 
                                                                                                                            720 
             C70                           966 912 811 849 
                                                                                                               806 
                                                                                                                            815 
             C90                           1035 948 913 839 787 
                                                                                                                            844 
             RPC200                        879 887 947 824     
                                                                                                                               
             RPC/AF                        978 948 974 1037 
                                                                                                               882 
                                                                                                                            746 
             CRC                           854 808 796 681 
                                                                                                               738 
                                                                                                                            757 


              Table 5.5  experimental data of specific heat for high and ultra high 
             performance concrete [51], * stands for dry 


In particular it is possible to note the difference between the value of 
specific heat obtained keeping the specimen at 105°C for ten days (indicated 
also as "dry") and the value obtained after only few hours of testing at the 
same temperature. 
Specific heat of C60 and C60-SF in dry state are comparable to those given 
in literature: around 800-900 J/kgK. Values in the saturated state are 
higher: 1000-1100 J/kgK. The water loss explain this decrease. 
As expected the specific heat of dry concrete increases with temperature 
because of deterioration of inner structure, i.e. for thermal dilatation and 
particularly for dehydration the porosity increases with temperature. This 
effect produces a material with a higher thermal inertia. 
According to the values of thermal conductivity for ultra-high performance 
concrete, these materials present a lower specific heat. This result is in part 
due to their higher compactness and in part due to the steel fibres 
distributed in the mass of cement. 
 
    

 231


                                                                                                              Chapter 5: Modelling of concrete at high temperature: thermodynamics, heat and mass transfer 




 

                                                                                                              1100                                                                                                                                                            1200
                                                                                                                                                          0.00E+00                                                                                                                                                      0.00E+00
                                                                                                              1050                 C60                    1.50E+08                                                                                                            1100                                      1.50E+08
                                                                                                                                                                                                                                                                                                   C60-SF
                                                                                                              1000
        ]                                                                                                                                                 3.00E+08         ]                                                                                                                                            3.00E+08
                                                                                                                                                          4.50E+08
             gK                                                                                                950                                                                                                                                                                                                      4.50E+08
                                                                                                                                                                                gK 1000
                   k                                                                                                                                      6.00E+08              k                                                                                                                                       6.00E+08
                                                                                                               900                                        Exper. values                                                                                                        900                                      Exper. values
                                                                                                               850
                        c heat [J/                                                                             800
                                      fi                                                                                                                                             c heat [J/                                                                                800
                                                                                                                                                                                                   fi
                                            eci                                                                750
                                                   p                                                                                                                                                     ecip 700
                                                        S                                                      700                                                                                       S
                                                                                                               650                                                                                                                                                             600

                                                                                                               600                                                                                                                                                             500
                                                                                                                      20    120    220      320    420          520                                                                                                                   20    120    220       320    420          520
                                                                                                                                   Temperature [°C]                                                                                                                                                 Temperature [°C]

                                                                                                              1100                                                                                                                                                            900
                                                                                                                                                       0.00E+00                                                                                                                                                            0.00E+00
                                                                                                              1050                 C90                 1.50E+08                                                                                                                                                            1.50E+08
                                                                                                                                                                                                                                                                              850                  CRC
                                                                                                              1000                                     3.00E+08                                                                                                                                                            3.00E+08
                                                             ]                                                                                                                                                ]
                                                                                                                                                       4.50E+08                                                    K                                                                                                       4.50E+08
                                                                  gK                                           950                                                                                                      g
                                                                        k                                                                              6.00E+08                                                                                                               800
                                                                                                                                                                                                                             /k                                                                                            6.00E+08
                                                                                                               900                                                                                                                 J
                                                                                                                                                       Eper. values                                                                     t [                                                                                Exper. values
                                                                                                                                                                                                                                               a
                                                                                                               850                                                                                                                                  e 750
                                                                                                                                                                                                                                                     h
                                                                             c heat [J/                        800
                                                                                           fi                                                                                                                                                             icifc
                                                                                                 eci                                                                                                                                                                          700
                                                                                                               750                                                                                                                                                 e
                                                                                                        pS                                                                                                                                                              Sp
                                                                                                               700                                                                                                                                                            650
                                                                                                               650

                                                                                                               600                                                                                                                                                            600
                                                                                                                      20    120    220      320    420          520                                                                                                                  20     120    220       320    420          520
                                                                                                                                   Temperature [°C]                                                                                                                                                Temperature [°C]



                                                                                                                       Figure 5.44  specific heat as direct function of temperature and capillary 
                                                                                                                       pressure for high and ultra high performance concrete provided in 
                                                                                                                       "Hiteco" project 
 
    




 232


  Concrete at high temperature: heat and mass transfer, inner structure and thermal properties 



