
172 The state equation of liquid water is rather complex [2],[13] and usually creates numerical problems. As a sufficient approximation of the dependence of the bulk density of liquid water upon its temperature and pressure, the following formula may be used [13],[14], ( ) ( ) 1 w wo w wo w o wp p Á Á ² ¸ ¸ ± î ù =     +   ï ú ð û (5.3) where Áwo= 999.84 kg/m3 is the water density at the reference temperature qo and pressure pwo, ( ) 1 w w p  ¸ = is volume thermal expansion coefficient of the water, 1 w v v p ¸   ( ±w=4.3Å10-9 Pa-1 [14]), and ( ) 1 w w v Á   = is the water specific volume. The coefficient ²w changes in non-linear manner with temperature (e.g. ²w=0.68Å10-4K-1 at q=273.15 K and ²w=10.1Å10-4K-1 at q=420 K [13]), thus the average value for the temperature range under interest should be used in calculations. The state equation (5.3), basically valid for the bulk (free) liquid water, is often used also for the description of the capillary and bound water, e.g. [15]-[18]. However, because of a very complex nature of interaction between the water and skeleton, its applicability is questionable, especially as far as pressure dependence is concerned (although capillary water is in traction, but its density is not expected to be lower than that for bulk water). Nevertheless, equation (5.3) may be used, taking into account that close to
Physics and thermodynamics of fluids in concrete at high temperature 173 critical point of water a sharp increase of volume thermal expansion coefficient takes place and assuming water incompressibility, i.e. ±w=0. In Figure 5.3 a linear relationship with ²w=constant and ±w=4.3Å10-9 Pa-1 is shown. This law is valid up to 200°C. Figure 5.3 Water density: linear relationship eq. (5.3) with ²w=constant, ±w=4.3Å10-9 Pa-1 Another relationship that can be used as state equation of water is a nonlinear expression of empirical nature due to Furbish [19]. Also in this case the incompressibility of water can be supposed. By assuming that w v remains constant, i.e. the liquid water compressibility tends to zero, 0 w ±   : 1 1 w w w w w w w H H v p v p Á ± Á     = =      (5.4) where w v is the molar volume of water and H is the main curvature of the liquid-vapour interface : ( ) ( )( ) 2 3 4 5 5 0 1 2 3 4 2 3 4 5 5 0 1 1 2 3 4 w w wrif b b b b b b p p a a a a a a Á ¸ ¸ ¸ ¸ ¸ ¸ ¸ ¸ ¸ ¸ = + + + + + +   + + + + + (5.5) where: pw1=1e07 Pa and pwrif=2e07 Pa; the coefficients a and b in equation (5.5) are shown in table 5.1: 900 920 940 960 980 1000 1020 273 373 473 573 673 Temperature [K] Liquid water density [kg/m3] 100 kPa 200 kPa 500 kPa 1000 kPa
174 a0 a1 a2 a3 a4 a5 4.89e07 -1.65e-09 1.86e-12 2.43e-13 -1.60e-15 3.37e-18 b0 b1 b2 b3 b4 b5 1.02e-03 -7-74e-01 8.77e-03 -9.21e-05 3.35e-07 -4.40e-10 Table 5.1 coefficients of equation (5.4) The relationship of eq. (5.5) is reported in Figure 5.4 where is evident the dependence of water density on temperature and the thermal expansion of water close to critical point. Figure 5.4 Water density: comparison between Furbish formula (5.5) and experimental values 5.2.2 Saturated vapour pressure The saturated water vapour pressure, i.e. the vapour pressure with flat surface (H=0), can be calculated through the well known Clasius-Clapeyron equation (see Chapter 2). Anyway, this relationship should be used carefully over extended temperature intervals, because enthalpy of vaporization is not constant but it is a function of temperature, i.e. ( ) vap vap h h ¸   =   and specific volumes vw and vv, of liquid water and water vapour respectively, change with temperature, too (see Fig. 5.6 and 5.7). More accurate results can be obtained using empirical correlations such as Hyland-Wexler relationship (Figure 5.5): 0 200 400 600 800 1000 1200 0 100 200 300 400 Temperature [°C] Liquid water density [kg/m3] Furbish formula Exper. values
175 1 2 3 4 6 exp / ln gwsp where C1=5800.2206, C2=1.3914993, C3=-4.8640293.10-2, C4=4.1764768.10-5, C5=-1.4452093.10-8, C6=6.5459673. Figure 5.5 Saturation vapour pressure as function of temperature (Hyland-Wexler formula) in comparison with experimental values 5.2.3 Thermodynamic equilibrium relationships It is well known that due to the curvature of the interface between the capillary- or bound-water and the gas phase inside the pores of the medium (considered as a capillary porous body), the equilibrium water vapour pressure pgw differs from the saturation pressure pgws and can be obtained from the Kelvin equation [20], ( ) ( ) exp c w gw gws gws w p M p p RH p R ¸ ¸ Á ¸ ë ö÷ ì ÷ = Å =   ì ÷ ì ÷ ìí ø (5.7) where RH is relative humidity, and the water vapour saturation pressure pgws, depending only upon temperature q, can be calculated from the Clausius-Clapeyron equation (see Chapter 2) or from empirical correlations, e.g. the formula proposed by Hyland and Wexler (5.6). 0,00E+00 2,00E+05 4,00E+05 6,00E+05 8,00E+05 1,00E+06 1,20E+06 1,40E+06 1,60E+06 1,80E+06 275 285 295 305 315 325 335 345 355 365 375 385 395 405 415 425 435 445 455 465 Temperature [K] Hyland-Wexler Experimental values
176 The radius of the curvature of the interface, Rc, being in the mechanical equilibrium state with water vapour, and corresponding to it pore radius, r, are given by the Young-Laplace equation [15],[20], wa c c p R Ã =, (5.8) where for spherical meniscus (capillary water) its curvature equals to 1 2 cos c a R r t Ñ Å =   (5.9) while for cylindrical interface (bound water) it is 1 1 R r t c a =   (5.10) Ãwa is the surface tension of water-air interface, J is the wetting angle, usually assumed to be zero, hence cos J=1, and ta is the thickness of the adsorbed water layer in the pore (see Fig. 5.20). As briefly mentioned in Chapter 2, it should be emphasised that equations (5.7) and (5.8) are valid both for the capillary and adsorbed water [20]. In the latter case the capillary pressure pc should be understood as the matric potential ¦, being a thermodynamic potential referring to the bound water, which cannot be identified to the pressure [15] as it will be shown below. For a mixture of N components, the equilibrium between a liquid (l) and its vapour (v) can be expressed using the chemical potential µ as (supposing that chemical equilibrium exists): 1,2,......., i i v l i N µ µ= = (5.11) or, since for a pure substance the chemical potential µ is equal to the molar Gibbs function g of each phase, the equilibrium can also be expressed: v lg g = (5.12) The differential of the chemical potential for a single phase of pure substance can be expressed as: dg d v dp dx dLÃ » ¸ µ =   + +   (5.13) where » is the molar entropy, v the molar volume, x the molar fraction and dL Ã the amount of work made by the surface tension to extend the phase separation surface. The phase interface is assumed to have a constant area
Physics and thermodynamics of fluids in concrete at high temperature 177 (i.e. dL Ã=0) and for a pure substance the molar fraction x is always constant, then the variation of the molar Gibbs function is: dg d v dp » ¸ =   + (5.14) It is important to highlight that the interfacial layer is not included in the analysis that means that the interfacial volume is negligible, i.e. 0 t V   , hence the Helmotz free energy of the interface tends to zero because there is only the work of surface formation whose variation is zero (as above mentioned, the interface is assumed to have a constant area), and the complete and ordinary chemical potential are equal (the derivatives of Helmotz free energy of the interfacial surface on numbers of moles are zero). Another hypothesis is that the meniscus mean radius of curvature is larger than the interfacial thickness, i.e. H-1> 100ú otherwise the effect of curvature on surface tension is not more negligible (H is the curvature of the interface). Since v l d d µ µ = and thermodynamic equilibrium requires thermal equilibrium, i.e., equal temperature in the two phases v l d d d ¸ ¸ ¸ = = ), we can write: v v v l l l d vdp d vdp » ¸ » ¸   + =  + (5.15) As state equations for water vapour dry air and their mixture, i.e., moist air, one can use the state equation for ideal gas and Dalton s law as explained in Chapter 2. These relationships are, however, valid when the range of pressures and temperatures of the analysed process is close to the ambient ones. Derivation of the Kelvin equation The Kelvin s formula can be obtained remembering that the capillary pressure pc is defined as: 2 c v l p p p HÃ =   = (5.16) in fact, in the following analysis the effect of meniscus curvature also appears through 2H Ã, which is the same as the capillary pressure because we suppose that the mechanical equilibrium exists, i.e., the Young-Laplace equation (eq.(5.8)) is valid. In differential form: ( ) 2 c v l dp dp dp d HÃ =   = (5.17) The molar based Gibbs-Duhem equation (5.13) for the liquid l and its vapour v phase becomes:
178 0 0 l l l l v v v v d vdp d d vdp d » ¸ µ » ¸ µ   + =   + = (5.18) and, since there is chemical equilibrium (5.11) i.e. v l d d d µ µ µ = = and considering a constant temperature process ( 0 d¸ = ) eq. (5.18) becomes: constant v v l l v dp v dp ¸ = = (5.19) and substituting (5.17) in (5.19) to eliminate pl: ( ) 2 v v l c v v v l l v v v d H dp dp dp dp v v Ã   = = =   (5.20) where as we mentioned Ã is a weak function of H. By assuming a perfect gas behaviour for the vapour, i.e.: 1 or v v v v w w v R p p M M R v Á ¸ Á ¸ = =a (5.21) we have: ( ) 2 , constant c v v v l R d H dp dp dp v p ¸ Ã ¸ = =   = (5.22) which can be integrated from pv=pv0 and pc=0 for H0=0 (i.e. flat surface) to pv and pc for H`0. By assuming that l v remains constant, i.e. the liquid water compressibility tends to zero, 0 l ±   we have: ( ) ( )0 0 2 ln , and are kept constant v c v v l v l R p H p p p v v p ¸ Ã ¸ ë ö÷ ì ÷ = =     ì ÷ ì ÷ í ø (5.23) This is the Kelvin equation and shows that for the case considered here the vapour pressure decreases as H (curvature of meniscus) increases (or the mean radius decreases). Usually the first term on the right hand side of (5.23) is negligible. In fact, for water at 18°C and a mean radius of curvature of 1µm, we have pv/pv0=0.9891, and for a mean radius of curvature of 100ú, this becomes 0.897. If we allow for the change of l v with the pressure, a correction can be added to (5.23). However, this correction is generally small. To show this, we integrate (5.4) to arrive at: 0 ln ( ) l l l l lo v p p v ± =    (5.24)
179 Now we note that for water l ± =5 10-11 dyne/cm2, and therefore the change in l v is negligible. Other consequences of effect of curvature in a single component system are the reduction of chemical potential, the increase of heat of evaporation and the change in triple point temperature. Clapeyron equation and Kelvin equation are strictly valid only for a pure vapour. For the case of a liquid in equilibrium with a mixture of ideal gases having a total pressure pg, to which is applicable Dalton s law g gw ga p p p = + , instead of equation (5.23) we obtain: ( ) ( )0 2 ln gw c g g w gws R p H p p p v p ¸ Ã ë ö÷ ì = =     ÷ ì ÷ í ø (5.25) For equilibrium between water and moist air, the total gas pressure can be usually considered constant (pg=pg0=atmospheric pressure). Moreover, in ordinary situations (pgªatmospheric pressure, i.e. about 1 bar) since the molar volume of the liquid phase w v is much smaller than the specific volume of its vapour gw v , the influence of the total gas pressure pg on the vapour tension of the liquid pgws can be neglected. In such a case, equation (5.23) can be used, but now pv has the meaning of the partial vapour pgw pressure in the mixture and pv0 of the vapour tension of the liquid pgws (pgws = pgw for a flat surface, when pc=0). These hypothesis are surely valid in the normal range of variation of pressure and temperature, i.e. far away the critical point of water. But analysing the behaviour of concrete at high temperature it is clear that the situation is different. When temperature reaches values close to the critical point of water, an additional increase of water degree of saturation, caused by considerable thermal dilatation of the liquid water, is observed. The last phenomena cause an effect, referred sometimes as  moisture clog , and are a reason of existence of the four zones, where moisture content is higher than the initial one, and where even fully saturated conditions may be sometimes encountered. In particular the water vapour does not behave as a ideal gas in the range of temperature between 160°C and the critical point of water. If we consider the vapour as a perfect gas for calculating its density from the saturated pressure, i.e., in flat conditions as far as the interface is concerned, we obtain values two-three times lower than real values available from thermodynamic tables, see Figure 5.6-5.7. Considering the correct variation of molar volume or density with temperature, e.g. to define gw v or Ágw as mainly function of temperature, one obtains: ( ) ( ) ref ref , for , for gw gw gw gw gw crit w gw gw gw gw gw crit R v v p v p M p p R ¸ ¸ ¸ ¸¸ Á Á ¸ Á ¸ ¸ ¸ ¸ î ù = = +  < < ï ú ï ú ð û î ù = = +  < < ï ú ð û (5.26)
180 Where qref is the reference temperature (around 160°C) the additive terms gw v   and gw Á   are exclusively function of temperature taking into account the real behaviour of water vapour in the range 160-373.15°C (Figure 5.6- 5.7). The firsts terms on the right hand side of eq. (5.26) are the terms related to ideal behaviour. Figure 5.6 Water vapour and liquid water molar volume: experimental data Figure 5.7 Comparison of the experimental values for water vapour density with those calculated by using (5.26) and by using ideal gas hypothesis 0.E+00 5.E+01 1.E+02 2.E+02 2.E+02 3.E+02 3.E+02 4.E+02 0 50 100 150 200 250 300 350 400 Temperature [°C] Vapour density [ kg/m3] Exper.values Ideal Gas Numerical approx. 0.00001 0.0001 0.001 0.01 0.1 1 10 273.15 323.15 373.15 423.15 473.15 523.15 573.15 623.15 673.15 Temperature [K] Molar volume [m3/mol] Water vapour Liquid water
181 Energetic meaning of capillary pressure Above critical point of water, the liquid water and water vapour do not exist and there is only the gas phase. As a result, there are no phase changes of the pore water (condensation   evaporation) and capillary pressure has not any physical meaning. Thus it cannot be directly used as a state variable for water, as for example vapour pressure does. Hence in these conditions we deal with a kind of the Stefan problem, where two regions of concrete, with the temperature below and above the critical point of water, are separated by the moving boundary. In these two regions the hygro-thermal state of the material is described by the different sets of state variables and equations expressing conservation laws of mass and energy for the system. There are some numerical methods for dealing with the Stefan problem, e.g., but they are numerically costly in the case of 2-D, and especially, 3-D problems. In this chapter a method which formally allows us to avoid direct tracing of the boundary position in the space will be presented. We give different physical meaning to the capillary pressure in the zone, where temperature exceeds the critical point of water, and apply a special  switching  procedure of the finite element from the below- to the above critical temperature description of the medium. Above the critical point of water the moisture may be present in the pores of concrete in two forms, as the adsorbed, liquid water and the gaseous water. The latter forms a mixture with the dry air and is assumed to behave as an ideal gas. We assume that the adsorbed water is in thermodynamic equilibrium with the gaseous phase of water, thus it may be described by the equation [21],[22]: ln gw gws p R p ¸ ë ö ¨ = ì ÷ í ø (5.27) where ¨ = -  Hws / Mw is the so called water potential,  Hws   the enthalpy of water adsorption by the concrete skeleton, pgw   partial vapour pressure and pgws   equilibrium vapour pressure at a flat adsorbate surface). Below the critical point of water, Tcr, the value of the water potential is related to capillary pressure according to: pc = - ¨ Å Áw (5.28) where Áw is the liquid water density. The surface diffusion of the adsorbed water on the solid skeleton may be described by: w bD grad =   ¨ j (5.29) with jw being mass flux of the adsorbed water.
182 For the gaseous water there are two transport mechanisms: advection and diffusion, described by a similar formula as the one for the situation below the critical temperature of water. Using equation (5.27), one may substitute in previous equations vapour pressure and its gradient with the appropriate expressions containing water potential ¨. The latter is in fact the state variable which describes thermodynamic state of both the adsorbed water and the water vapour. Because of lack of any trustable data concerning the equilibrium vapour pressure pgws above temperature Tcr, we assume that it follows the perfect gas law, i.e. increases linearly with temperature. For the same reason we assume that the density of adsorbed water above the critical temperature Tcr is equal to the water vapour density, which is equal to the water density at the critical point (Fig.5.5-5.6). Now, using (5.28), we introduce formally into equations describing mass and energy transport the capillary pressure to obtain the same form of these equations as below the temperature Tcr. One must however remember that the capillary pressure only substitutes formally the water potential ¨ and they have different physical interpretation. Because of different physical meaning of the capillary pressure pc and different nature of the physical phenomena above the temperature Tcr (there is not liquid water   vapour phase change and associated with it heat effect), we have introduced a special  switching  procedure, which is applied to the whole finite elements. When part of FE nodes have temperature above Tcr, the capillary pressure is blocked at the previous value, until the temperature in all the nodes does pass the critical point of water. 5.2.4 Enthalpy of vaporization The enthalpy of evaporation depends upon the temperature and may be approximated by the Watson formula, [13],[14]: ( )0.38 5 2.672 10 vap cr h ¸ ¸   = Å   (5.30) where qcr= 647.3 K is the critical temperature of water. A comparison of the experimental data from [23],[24] and those calculated by using equation (5.30) is shown in Figure 5.8. 5.2.5 Viscosity of moist air The dynamic viscosity of moist air, which depends upon the temperature and the ratio of the vapour and gas pressures, using the data from [23],[25], can be approximated using the following formula [26]: 0.608 +( - ) ga g gw ga gwg p p µ µ µ µ ë ö÷ ì = ÷ ì ÷ í ø (5.31)
183 where ( ) 1 gw g p ( ) 0 0 with µgw0= 8.85Å10-6 PaÅs, ±v=3.53Å10-8 PaÅsÅK-1, ( ) ( )2 0 0 0 with µga0= 17.17Å10-6 PaÅs, ±a= 4.73Å10-8 PaÅsÅK-1, ²a= 2.22Å10-11 PaÅsÅK-2. A comparison of the experimental data of Mason and Monchick [45] for moist air with those calculated using equation (5.31) is shown in Figure 5.9. 5.2.6 Viscosity of liquid water The dynamic viscosity of liquid water depends strongly upon temperature [13] and can be evaluated, with sufficient accuracy in a wide temperature range, using the following approximate formula, [27] ( )1.562 0.6612 229 w T µ   = Å   (5.32) which results are compared with the experimental data [13],[24] in Figure Figure 5.8 Comparison of the experimental values for evaporation enthalpy with those calculated by using the Watson formula (5.30) 0.00E+00 5.00E+05 1.00E+06 1.50E+06 2.00E+06 2.50E+06 3.00E+06 273 323 373 423 473 523 573 623 673 Temperature [K] Watson formula Exper. values
Figure 5.9 Comparison of the experimental data for the dynamic viscosity of moist air with those calculated using approximated formula (5.31)
Figure 5.10 Comparison of the experimental data for the dynamic viscosity of water with those calculated with Thomas formula (5.32)
